\chap{Formal Grammars}
The following two listings constitute the formal grammar of
    Lojban. The first version is written in the YACC language,
    which is used to describe parsers, and has been used to create
    a parser for Lojban texts. This parser is available from the
    Logical Language Group. The second listing is in Extended
    Backus-Naur Form (EBNF) and represents the same grammar in a
    more human-readable form. (In case of discrepancies, the YACC
    version is official.) There is a cross-reference listing for
    the EBNF format that shows, for each selma'o and rule, which
    rules refer to it.



\sect{YACC Grammar of Lojban}
<div class="c2">
      /* LOJBAN MACHINE GRAMMAR, 3RD BASELINE AS OF 10 JANUARY
      1997<br />
       WHICH IS ORIGINAL BASELINE 20 JULY 1990 INCORPORATING JC'S
      TECH FIXES 1-28<br />
       THIS DRAFT ALSO INCORPORATES CHANGE PROPOSALS 1-47 DATED 29
      DECEMBER 1996<br />
      <br />
       THIS DOCUMENT IS EXPLICITLY DEDICATED TO THE PUBLIC
      DOMAIN<br />
       BY ITS AUTHOR, THE LOGICAL LANGUAGE GROUP INC.<br />
       CONTACT THAT ORGANIZATION AT 2904 BEAU LANE, FAIRFAX VA
      22031 USA<br />
       U.S. PHONE: 703-385-0273<br />
       INTL PHONE: +1 703 385-0273<br />
    </div>
    <br />
    <br />
     grammar.300 */ 

/* The Lojban machine parsing algorithm is a multi-step
    process. The YACC machine grammar presented here is an amalgam
    of those steps, concatenated so as to allow YACC to verify the
    syntactic ambiguity of the grammar. YACC is used to generate a
    parser for a portion of the grammar, which is LALR1 (the type
    of grammar that YACC is designed to identify and process
    successfully), but most of the rest of the grammar must be
    parsed using some language-coded processing.

<h4>Step 1 - Lexing</h4>

From phonemes, stress, and pause, it is possible to resolve
    Lojban unambiguously into a stream of words. Any machine
    processing of speech will have to have some way to deal with
    'non-Lojban' failures of fluent speech, of course. The resolved
    words can be expressed as a text file, using Lojban's phonetic
    spelling rules.

The following steps, assume that there is the possibility of
    non-Lojban text within the Lojban text (delimited
    appropriately). Such non-Lojban text may not be reducible from
    speech phonetically. However, step 2 allows the filtering of a
    phonetically transcribed text stream, to recognize such
    portions of non-Lojban text where properly delimited, without
    interference with the parsing algorithm.

<h4>Step 2 - Filtering</h4>

From start to end, performing the following filtering and
    lexing tasks using the given order of precedence in case of
    conflict:

a. If the Lojban word \q{zoi} (selma'o ZOI) is identified,
    take the following Lojban word (which should be end delimited
    with a pause for separation from the following non-Lojban text)
    as an opening delimiter. Treat all text following that
    delimiter, until that delimiter recurs *after a pause*, as
    grammatically a single token (labelled '\hyperref[html:y699]{anything\_699}' in this grammar). There is no
    need for processing within this text except as necessary to
    find the closing delimiter.

b. If the Lojban word \q{zo} (selma'o ZO) is identified,
    treat the following Lojban word as a token labelled '\hyperref[html:y698]{any\_word\_698}', instead of lexing it by its
    normal grammatical function.

c. If the Lojban word \q{lo'u} (selma'o LOhU) is identified,
    search for the closing delimiter \q{le'u} (selma'o LEhU),
    ignoring any such closing delimiters absorbed by the previous
    two steps. The text between the delimiters should be treated as
    the single token '\hyperref[html:y697]{any\_words\_697}'.

d. Categorize all remaining words into their Lojban selma'o
    category, including the various delimiters mentioned in the
    previous steps. In all steps after step 2, only the selma'o
    token type is significant for each word.

e. If the word \q{si} (selma'o SI) is identified, erase it
    and the previous word (or token, if the previous text has been
    condensed into a single token by one of the above rules).

f. If the word \q{sa} (selma'o SA) is identified, erase it
    and all preceding text as far back as necessary to make what
    follows attach to what precedes. (This rule is hard to
    formalize and may receive further definition later.)

g. If the word 'su' (selma'o SU) is identified, erase it and
    all preceding text back to and including the first preceding
    token word which is in one of the selma'o: NIhO, LU, TUhE, and
    TO. However, if speaker identification is available, a SU shall
    only erase to the beginning of a speaker's discourse, unless it
    occurs at the beginning of a speaker's discourse. (Thus, if the
    speaker has said something, two \q{su}'s are required to erase
    the entire conversation.

<h4>Step 3 - Termination</h4>

If the text contains a FAhO, treat that as the end-of-text
    and ignore everything that follows it.

<h4>Step 4 - Absorption of Grammar-Free Tokens</h4>

In a new pass, perform the following absorptions (absorption
    means that the token is removed from the grammar for processing
    in following steps, and optionally reinserted, grouped with the
    absorbing token after parsing is completed).

a. Token sequences of the form any - (ZEI - any) ..., where
    there may be any number of ZEIs, are merged into a single token
    of selma'o BRIVLA.

b. Absorb all selma'o BAhE tokens into the following token.
    If they occur at the end of text, leave them alone (they are
    errors).

c. Absorb all selma'o BU tokens into the previous token.
    Relabel the previous token as selma'o BY.

d. If selma'o NAI occurs immediately following any of tokens
    UI or CAI, absorb the NAI into the previous token.

e. Absorb all members of selma'o DAhO, FUhO, FUhE, UI, Y,
    and CAI into the previous token. All of these null grammar
    tokens are permitted following any word of the grammar, without
    interfering with that word's grammatical function, or causing
    any effect on the grammatical interpretation of any other token
    in the text. Indicators at the beginning of text are explicitly
    handled by the grammar.

<h4>Step 5 - Insertion of Lexer Lexemes</h4>

Lojban is not in itself LALR1. There are words whose
    grammatical function is determined by following tokens. As a
    result, parsing of the YACC grammar must take place in two
    steps. In the first step, certain strings of tokens with
    defined grammars are identified, and either

a. are replaced by a single specified 'lexer token' for step
    6, or

b. the lexer token is inserted in front of the token string
    to identify it uniquely.

The YACC grammar included herein is written to make YACC
    generation of a step 6 parser easy regardless of whether a. or
    b. is used. The strings of tokens to be labelled with lexer
    tokens are found in rule terminals labelled with numbers
    between 900 and 1099. These rules are defined with the lexer
    tokens inserted, with the result that it can be verified that
    the language is LALR1 under option b. after steps 1 through 4
    have been performed. Alternatively, if option a. is to be used,
    these rules are commented out, and the rule terminals labelled
    from 800 to 900 refer to the lexer tokens *without* the strings
    of defining tokens. Two sets of lexer tokens are defined in the
    token set so as to be compatible with either option.

In this step, the strings must be labelled with the
    appropriate lexer tokens. Order of inserting lexer tokens *IS*
    significant, since some shorter strings that would be marked
    with a lexer token may be found inside longer strings. If the
    tokens are inserted before or in place of the shorter strings,
    the longer strings cannot be identified.

If option a. is chosen, the following order of insertion
    works correctly (it is not the only possible order): A, C, D,
    B, U, E, H, I, J, K, M ,N, G, O, V, W, F, P, R, T, S, Y, L, Q.
    This ensures that the longest rules will be processed first; a
    PA+MAI will not be seen as a PA with a dangling MAI at the end,
    for example.

<h4>Step 6 - YACC Parsing</h4>

YACC should now be able to parse the Lojban text in
    accordance with the rule terminals labelled from 1 to 899 under
    option 5a, or 1 to 1099 under option 5b. Comment out the rules
    beyond 900 if option 5a is used, and comment out the 700-series
    of lexer-tokens, while restoring the series of lexer tokens
    numbered from 900 up.

*/

%token \phantomsection\label{html:y501}
A_501            /*        eks; basic afterthought logical connectives */
%token \phantomsection\label{html:y502}
BAI_502          /*        modal operators */
%token \phantomsection\label{html:y503}
BAhE_503         /*        next word intensifier */
%token \phantomsection\label{html:y504}
BE_504           /*        sumti link to attach sumti to a selbri */
%token \phantomsection\label{html:y505}
BEI_505          /*        multiple sumti separator between BE, BEI */
%token \phantomsection\label{html:y506}
BEhO_506         /*        terminates BE/BEI specified descriptors */
%token \phantomsection\label{html:y507}
BIhI_507         /*        interval component of JOI */
%token \phantomsection\label{html:y508}
BO_508           /*        joins two units with shortest scope */
%token \phantomsection\label{html:y509}
BRIVLA_509       /*        any brivla */
%token \phantomsection\label{html:y511}
BU_511           /*        turns any word into a BY lerfu word */
%token \phantomsection\label{html:y513}
BY_513           /*        individual lerfu words */
%token \phantomsection\label{html:y514}
CAhA_514         /*        specifies actuality/potentiality of tense */
%token \phantomsection\label{html:y515}
CAI_515          /*        afterthought intensity marker */
%token \phantomsection\label{html:y516}
CEI_516          /*        pro-bridi assignment operator */
%token \phantomsection\label{html:y517}
CEhE_517         /*        afterthought term list connective */
%token \phantomsection\label{html:y518}
CMENE_518        /*        names; require consonant end, then pause no
                                   LA or DOI selma'o embedded, pause before if
                                   vowel initial and preceded by a vowel */
%token \phantomsection\label{html:y519}
CO_519           /*        tanru inversion  */
%token \phantomsection\label{html:y520}
COI_520          /*        vocative marker permitted inside names; must
                                   always be followed by pause or DOI */
%token \phantomsection\label{html:y521}
CU_521           /*        separator between head sumti and selbri */
%token \phantomsection\label{html:y522}
CUhE_522         /*        tense/modal question */
%token \phantomsection\label{html:y524}
DAhO_524         /*        cancel anaphora/cataphora assignments */
%token \phantomsection\label{html:y525}
DOI_525          /*        vocative marker */
%token \phantomsection\label{html:y526}
DOhU_526         /*        terminator for DOI-marked vocatives */
%token \phantomsection\label{html:y527}
FA_527           /*        modifier head generic case tag */
%token \phantomsection\label{html:y528}
FAhA_528         /*        superdirections in space */
%token \phantomsection\label{html:y529}
FAhO_529         /*        normally elided 'done pause' to indicate end
                                   of utterance string */
%token \phantomsection\label{html:y530}
FEhE_530         /*        space interval mod flag */
%token \phantomsection\label{html:y531}
FEhU_531         /*        ends bridi to modal conversion */
%token \phantomsection\label{html:y532}
FIhO_532         /*        marks bridi to modal conversion */
%token \phantomsection\label{html:y533}
FOI_533          /*        end compound lerfu */
%token \phantomsection\label{html:y535}
FUhE_535         /*        open long scope for indicator */
%token \phantomsection\label{html:y536}
FUhO_536         /*        close long scope for indicator */
%token \phantomsection\label{html:y537}
GA_537           /*        geks; forethought logical connectives */
%token \phantomsection\label{html:y538}
GEhU_538         /*        marker ending GOI relative clauses */
%token \phantomsection\label{html:y539}
GI_539           /*        forethought medial marker */
%token \phantomsection\label{html:y541}
GIhA_541         /*        logical connectives for bridi-tails */
%token \phantomsection\label{html:y542}
GOI_542          /*        attaches a sumti modifier to a sumti */
%token \phantomsection\label{html:y543}
GOhA_543         /*        pro-bridi */
%token \phantomsection\label{html:y544}
GUhA_544         /*        GEK for tanru units, corresponds to JEKs */
%token \phantomsection\label{html:y545}
I_545            /*        sentence link */
%token \phantomsection\label{html:y546}
JA_546           /*        jeks; logical connectives within tanru */
%token \phantomsection\label{html:y547}
JAI_547          /*        modal conversion flag */
%token \phantomsection\label{html:y548}
JOI_548          /*        non-logical connectives */
%token \phantomsection\label{html:y550}
KEhE_550         /*        right terminator for KE groups */
%token \phantomsection\label{html:y551}
KE_551           /*        left long scope marker */
%token \phantomsection\label{html:y552}
KEI_552          /*        right terminator, NU abstractions */
%token \phantomsection\label{html:y554}
KI_554           /*        multiple utterance scope for tenses */
%token \phantomsection\label{html:y555}
KOhA_555         /*        sumti anaphora */
%token \phantomsection\label{html:y556}
KU_556           /*        right terminator for descriptions, etc. */
%token \phantomsection\label{html:y557}
KUhO_557         /*        right terminator, NOI relative clauses */
%token \phantomsection\label{html:y558}
LA_558           /*        name descriptors */
%token \phantomsection\label{html:y559}
LAU_559          /*        lerfu prefixes */
%token \phantomsection\label{html:y561}
LAhE_561         /*        sumti qualifiers */
%token \phantomsection\label{html:y562}
LE_562           /*        sumti descriptors */
%token \phantomsection\label{html:y565}
LEhU_565         /*        possibly ungrammatical text right quote */
%token \phantomsection\label{html:y566}
LI_566           /*        convert number to sumti */
%token \phantomsection\label{html:y567}
LIhU_567         /*        grammatical text right quote */
%token \phantomsection\label{html:y568}
LOhO_568         /*        elidable terminator for LI */
%token \phantomsection\label{html:y569}
LOhU_569         /*        possibly ungrammatical text left quote */
%token \phantomsection\label{html:y571}
LU_571           /*        grammatical text left quote */
%token \phantomsection\label{html:y573}
LUhU_573         /*        LAhE close delimiter */
%token \phantomsection\label{html:y574}
ME_574           /*        converts a sumti into a tanru_unit */
%token \phantomsection\label{html:y575}
MEhU_575         /*        terminator for ME */
%token \phantomsection\label{html:y577}
MOhI_577         /*        motion tense marker */
%token \phantomsection\label{html:y578}
NA_578           /*        bridi negation  */
%token \phantomsection\label{html:y581}
NAI_581          /*        attached to words to negate them */
%token \phantomsection\label{html:y583}
NAhE_583         /*        scalar negation  */
%token \phantomsection\label{html:y584}
NIhO_584         /*        new paragraph; change of subject */
%token \phantomsection\label{html:y585}
NOI_585          /*        attaches a subordinate clause to a sumti */
%token \phantomsection\label{html:y586}
NU_586           /*        abstraction  */
%token \phantomsection\label{html:y587}
NUhI_587         /*        marks the start of a termset */
%token \phantomsection\label{html:y588}
NUhU_588         /*        marks the middle and end of a termset */
%token \phantomsection\label{html:y591}
PEhE_591         /*        afterthought termset connective prefix */
%token \phantomsection\label{html:y592}
PU_592           /*        directions in time */
%token \phantomsection\label{html:y593}
RAhO_593         /*        flag for modified interpretation of GOhI */
%token \phantomsection\label{html:y594}
ROI_594          /*        converts number to extensional tense */
%token \phantomsection\label{html:y595}
SA_595           /*        metalinguistic eraser to the beginning of
                                   the current utterance */
%token \phantomsection\label{html:y596}
SE_596           /*        conversions */
%token \phantomsection\label{html:y597}
SEI_597          /*        metalinguistic bridi insert marker */
%token \phantomsection\label{html:y598}
SEhU_598         /*        metalinguistic bridi end marker */
%token \phantomsection\label{html:y601}
SI_601           /*        metalinguistic single word eraser */
%token \phantomsection\label{html:y602}
SOI_602          /*        reciprocal sumti marker */
%token \phantomsection\label{html:y603}
SU_603           /*        metalinguistic eraser of the entire text */
%token \phantomsection\label{html:y604}
TAhE_604         /*        tense interval properties */
%token \phantomsection\label{html:y605}
TEI_605          /*        start compound lerfu */
%token \phantomsection\label{html:y606}
TO_606           /*        left discursive parenthesis */
%token \phantomsection\label{html:y607}
TOI_607          /*        right discursive parenthesis */
%token \phantomsection\label{html:y610}
TUhE_610         /*        multiple utterance scope mark */
%token \phantomsection\label{html:y611}
TUhU_611         /*        multiple utterance end scope mark */
%token \phantomsection\label{html:y612}
UI_612           /*        attitudinals, observationals, discursives */
%token \phantomsection\label{html:y613}
VA_613           /*        distance in space-time */
%token \phantomsection\label{html:y614}
VAU_614          /*        end simple bridi or bridi-tail */
%token \phantomsection\label{html:y615}
VEhA_615         /*        space-time interval size */
%token \phantomsection\label{html:y616}
VIhA_616         /*        space-time dimensionality marker */
%token \phantomsection\label{html:y617}
VUhO_617         /*        glue between logically connected sumti
                                    and relative clauses */
%token \phantomsection\label{html:y618}
XI_618           /*        subscripting operator */
%token \phantomsection\label{html:y619}
Y_619            /*        hesitation */
%token \phantomsection\label{html:y621}
ZAhO_621         /*        event properties - inchoative, etc. */
%token \phantomsection\label{html:y622}
ZEhA_622         /*        time interval size tense */
%token \phantomsection\label{html:y623}
ZEI_623          /*        lujvo glue */
%token \phantomsection\label{html:y624}
ZI_624           /*        time distance tense */
%token \phantomsection\label{html:y625}
ZIhE_625         /*        conjoins relative clauses */
%token \phantomsection\label{html:y626}
ZO_626           /*        single word metalinguistic quote marker */
%token \phantomsection\label{html:y627}
ZOI_627          /*        delimited quote marker */
%token \phantomsection\label{html:y628}
ZOhU_628         /*        prenex terminator (not elidable) */

%token \phantomsection\label{html:y650}
BIhE_650         /*        prefix for high-priority MEX operator */
%token \phantomsection\label{html:y651}
BOI_651          /*        number or lerfu-string terminator */
%token \phantomsection\label{html:y655}
FUhA_655         /*        reverse Polish flag */
%token \phantomsection\label{html:y656}
GAhO_656         /*        open/closed interval markers for BIhI */
%token \phantomsection\label{html:y657}
JOhI_657         /*        flags an array operand */
%token \phantomsection\label{html:y658}
KUhE_658         /*        MEX forethought delimiter */
%token \phantomsection\label{html:y661}
MAI_661          /*        change numbers to utterance ordinals */
%token \phantomsection\label{html:y662}
MAhO_662         /*        change MEX expressions to MEX operators */
%token \phantomsection\label{html:y663}
MOI_663          /*        change number to selbri */
%token \phantomsection\label{html:y664}
MOhE_664         /*        change sumti to operand, inverse of LI */
%token \phantomsection\label{html:y665}
NAhU_665         /*        change a selbri into an operator */
%token \phantomsection\label{html:y666}
NIhE_666         /*        change selbri to operand; inverse of MOI */
%token \phantomsection\label{html:y667}
NUhA_667         /*        change operator to selbri; inverse of MOhE */
%token \phantomsection\label{html:y672}
PA_672           /*        numbers and numeric punctuation */
%token \phantomsection\label{html:y673}
PEhO_673         /*        forethought (Polish) flag */
%token \phantomsection\label{html:y675}
TEhU_675         /*        closing gap for MEX constructs */
%token \phantomsection\label{html:y677}
VEI_677          /*        left MEX bracket */
%token \phantomsection\label{html:y678}
VEhO_678         /*        right MEX bracket */
%token \phantomsection\label{html:y679}
VUhU_679         /*        MEX operator */

%token any_words_697    /*        a string of lexable Lojban words */
%token any_word_698     /*        any single lexable Lojban words */
%token anything_699     /*        a possibly unlexable phoneme string */

/* The following tokens are the actual lexer tokens.  The _900 series
tokens are duplicates that allow limited testing of lexer rules in the
context of the total grammar.  They are used in the actual parser, where
the 900 series rules are found in the lexer.  */

%token lexer_\phantomsection\label{html:y701}
A_701      /*        flags a MAI utterance ordinal */
%token lexer_\phantomsection\label{html:y702}
B_702      /*        flags an EK unless EK_BO, EK_KE */
%token lexer_\phantomsection\label{html:y703}
C_703      /*        flags an EK_BO */
%token lexer_\phantomsection\label{html:y704}
D_704      /*        flags an EK_KE */
%token lexer_\phantomsection\label{html:y705}
E_705      /*        flags a JEK */
%token lexer_\phantomsection\label{html:y706}
F_706      /*        flags a JOIK */
%token lexer_\phantomsection\label{html:y707}
G_707      /*        flags a GEK */
%token lexer_\phantomsection\label{html:y708}
H_708      /*        flags a GUhEK */
%token lexer_\phantomsection\label{html:y709}
I_709      /*        flags a NAhE_BO */
%token lexer_\phantomsection\label{html:y710}
J_710      /*        flags a NA_KU */
%token lexer_\phantomsection\label{html:y711}
K_711      /*        flags an I_BO (option. JOIK/JEK lexer tags)*/
%token lexer_\phantomsection\label{html:y712}
L_712      /*        flags a PA, unless MAI (then lexer A) */
%token lexer_\phantomsection\label{html:y713}
M_713      /*        flags a GIhEK_BO */
%token lexer_\phantomsection\label{html:y714}
N_714      /*        flags a GIhEK_KE */
%token lexer_\phantomsection\label{html:y715}
O_715      /*        flags a modal operator BAI or compound */
%token lexer_\phantomsection\label{html:y716}
P_716      /*        flags a GIK */
%token lexer_\phantomsection\label{html:y717}
Q_717      /*        flags a lerfu_string unless MAI (then lexer_A)*/
%token lexer_\phantomsection\label{html:y718}
R_718      /*        flags a GIhEK, not BO or KE */
%token lexer_\phantomsection\label{html:y719}
S_719      /*        flags simple I */
%token lexer_\phantomsection\label{html:y720}
T_720      /*        flags I_JEK */
%token lexer_\phantomsection\label{html:y721}
U_721      /*        flags a JEK_BO */
%token lexer_\phantomsection\label{html:y722}
V_722      /*        flags a JOIK_BO */
%token lexer_\phantomsection\label{html:y723}
W_723      /*        flags a JOIK_KE */
/* %token lexer_\phantomsection\label{html:y724}
X_724   /* null */
%token lexer_\phantomsection\label{html:y725}
Y_725      /*        flags a PA_MOI */

/*%token lexer_A_905    /*        :  lexer_A_701  utt_ordinal_root_906 */
/*%token lexer_B_910    /*        :  lexer_B_702  EK_root_911 */
/*%token lexer_C_915    /*        :  lexer_C_703  EK_root_911  BO_508 */
/*%token lexer_D_916    /*        :  lexer_D_704  EK_root_911  KE_551 */
/*%token lexer_E_925    /*        :  lexer_E_705  JEK_root_926 */
/*%token lexer_F_930    /*        :  lexer_F_706  JOIK_root_931 */
/*%token lexer_G_935    /*        :  lexer_G_707  GA_537 */
/*%token lexer_H_940    /*        :  lexer_H_708  GUhA_544 */
/*%token lexer_I_945    /*        :  lexer_I_709  NAhE_583  BO_508 */
/*%token lexer_J_950    /*        :  lexer_J_710  NA_578  KU_556 */
/*%token lexer_K_955    /*        :  lexer_K_711  I_432  BO_508 */
/*%token lexer_L_960    /*        :  lexer_L_712  number_root_961 */
/*%token lexer_M_965    /*        :  lexer_M_713  GIhEK_root_991  BO_508 */
/*%token lexer_N_966    /*        :  lexer_N_714  GIhEK_root_991  KE_551 */
/*%token lexer_O_970    /*        :  lexer_O_715  simple_tense_modal_972 */
/*%token lexer_P_980    /*        :  lexer_P_716  GIK_root_981 */
/*%token lexer_Q_985    /*        :  lexer_Q_717  lerfu_string_root_986 */
/*%token lexer_R_990    /*        :  lexer_R_718  GIhEK_root_991 */
/*%token lexer_S_995    /*        :  lexer_S_719  I_545 */
/*%token  lexer_T_1000  /*        :  lexer_T_720  I_545  simple_JOIK_JEK_957 */
/*%token lexer_U_1005   /*        :  lexer_U_721  JEK_root_926  BO_508 */
/*%token lexer_V_1010   /*        :  lexer_V_722  JOIK_root_931  BO_508 */
/*%token lexer_W_1015   /*        :  lexer_W_723  JOIK_root_931  KE_551 */
/*%token lexer_X_1020   /* null */
/*%token lexer_Y_1025   /*        :  lexer_Y_725  number_root_961  MOI_663 */

%start \hyperref[html:y0]{text_0}

%%

\phantomsection\label{html:y0}
text_0                  :  \hyperref[html:y1]{text_A_1}
                        |  \hyperref[html:y411]{indicators_411}  \hyperref[html:y1]{text_A_1}
                        |  \hyperref[html:y32]{free_modifier_32}  \hyperref[html:y1]{text_A_1}
                        |  \hyperref[html:y404]{cmene_404}  \hyperref[html:y1]{text_A_1}
                        |  \hyperref[html:y411]{indicators_411}  \hyperref[html:y32]{free_modifier_32}  \hyperref[html:y1]{text_A_1}
                        |  \hyperref[html:y581]{NAI_581}  \hyperref[html:y0]{text_0}
                        ;

\phantomsection\label{html:y1}
text_A_1                :  \hyperref[html:y422]{JOIK_JEK_422}  \hyperref[html:y2]{text_B_2}
                        /* incomplete JOIK_JEK  without preceding I */
                        /* compare note on \hyperref[html:y10]{paragraph_10} */
                        |  \hyperref[html:y2]{text_B_2}
                        ;

\phantomsection\label{html:y2}
text_B_2                :  \hyperref[html:y819]{I_819}  \hyperref[html:y2]{text_B_2}
                        |  \hyperref[html:y820]{I_JEK_820}  \hyperref[html:y2]{text_B_2}
                        |  \hyperref[html:y811]{I_BO_811}  \hyperref[html:y2]{text_B_2}
                        |  \hyperref[html:y410]{para_mark_410}  \hyperref[html:y3]{text_C_3}
                        |  \hyperref[html:y3]{text_C_3}
                        ;

\phantomsection\label{html:y3}
text_C_3                :  \hyperref[html:y4]{paragraphs_4}
   /* Only indicators which follow certain selma'o:
    cmene, \hyperref[html:y607]{TOI_607}, \hyperref[html:y571]{LU_571}, and the lexer_K and lexer_S I_roots and compounds,
    and at the start of text(_0), will survive the lexer; all other valid ones
    will be absorbed.  The only strings for which indicators generate a
    potential ambiguity are those which contain NAI.  An indicator cannot be
    inserted in between a token and its negating NAI, else you can't tell
    whether it is the indicator or the original token being negated. */
                        |  /* empty */
  /* An empty text is legal; formerly this was handled by the explicit
   appearance of \hyperref[html:y529]{FAhO_529}, but this is now absorbed by the preparser. */
                        ;

\phantomsection\label{html:y4}
paragraphs_4            :  \hyperref[html:y10]{paragraph_10}
                        |  \hyperref[html:y10]{paragraph_10}  \hyperref[html:y410]{para_mark_410}  \hyperref[html:y4]{paragraphs_4}
                        ;

\phantomsection\label{html:y10}
paragraph_10            :  \hyperref[html:y11]{statement_11}
                        |  \hyperref[html:y20]{fragment_20}
                        |  \hyperref[html:y10]{paragraph_10}  \hyperref[html:y819]{I_819}  \hyperref[html:y11]{statement_11}
                        |  \hyperref[html:y10]{paragraph_10}  \hyperref[html:y819]{I_819}  \hyperref[html:y20]{fragment_20}
                        |  \hyperref[html:y10]{paragraph_10}  \hyperref[html:y819]{I_819}
                        /* this last fixes an erroneous start to a sentence,
                           and permits incomplete JOIK_JEK after I, as well
                           in answer to questions on those connectives  */
                        ;

\phantomsection\label{html:y11}
statement_11            :  \hyperref[html:y12]{statement_A_12}
                        |  \hyperref[html:y30]{prenex_30}  \hyperref[html:y11]{statement_11}
                        ;

\phantomsection\label{html:y12}
statement_A_12          :  \hyperref[html:y13]{statement_B_13}
                        |  \hyperref[html:y12]{statement_A_12}  \hyperref[html:y820]{I_JEK_820}  \hyperref[html:y13]{statement_B_13}
                        |  \hyperref[html:y12]{statement_A_12}  \hyperref[html:y820]{I_JEK_820} 
                        ;

\phantomsection\label{html:y13}
statement_B_13          :  \hyperref[html:y14]{statement_C_14}
                        |  \hyperref[html:y14]{statement_C_14}  \hyperref[html:y811]{I_BO_811}  \hyperref[html:y13]{statement_B_13}
                        |  \hyperref[html:y14]{statement_C_14}  \hyperref[html:y811]{I_BO_811} 
                        ;

\phantomsection\label{html:y14}
statement_C_14          :  \hyperref[html:y40]{sentence_40}
                        |  \hyperref[html:y447]{TUhE_447}  \hyperref[html:y2]{text_B_2}  \hyperref[html:y454]{TUhU_gap_454}
                        |  \hyperref[html:y491]{tag_491}  \hyperref[html:y447]{TUhE_447}  \hyperref[html:y2]{text_B_2}  \hyperref[html:y454]{TUhU_gap_454}
                        ;

\phantomsection\label{html:y20}
fragment_20             :  \hyperref[html:y802]{EK_802}
                        |  \hyperref[html:y445]{NA_445}
                        |  \hyperref[html:y818]{GIhEK_818}
                        |  \hyperref[html:y300]{quantifier_300}
                        |  \hyperref[html:y80]{terms_80}  \hyperref[html:y456]{VAU_gap_456}        /* answer to ma */
                        /* \hyperref[html:y490]{mod_head_490} requires both \hyperref[html:y450]{gap_450} and \hyperref[html:y456]{VAU_gap_456}
                           but needs no extra rule to accomplish this */
                        |  \hyperref[html:y121]{relative_clauses_121}
                        |  \hyperref[html:y161]{links_161}
                        |  \hyperref[html:y160]{linkargs_160}
                        |  \hyperref[html:y30]{prenex_30}
                        ;

\phantomsection\label{html:y30}
prenex_30               :  \hyperref[html:y80]{terms_80}  \hyperref[html:y492]{ZOhU_492}
                        ;

\phantomsection\label{html:y32}
free_modifier_32        :  \hyperref[html:y33]{free_modifier_A_33}
                        |  \hyperref[html:y33]{free_modifier_A_33}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y33}
free_modifier_A_33      :  \hyperref[html:y35]{vocative_35}
                        |  \hyperref[html:y36]{parenthetical_36}
                        |  \hyperref[html:y34]{discursive_bridi_34}
                        |  \hyperref[html:y486]{subscript_486}
                        |  \hyperref[html:y801]{utterance_ordinal_801}
                        ;

\phantomsection\label{html:y34}
discursive_bridi_34     :  \hyperref[html:y440]{SEI_440}  \hyperref[html:y130]{selbri_130}  \hyperref[html:y459]{SEhU_gap_459}
                        |  \hyperref[html:y498]{SOI_498}  \hyperref[html:y90]{sumti_90}  \hyperref[html:y459]{SEhU_gap_459}
                        |  \hyperref[html:y498]{SOI_498}  \hyperref[html:y90]{sumti_90}  \hyperref[html:y90]{sumti_90}  \hyperref[html:y459]{SEhU_gap_459}
                        |  \hyperref[html:y440]{SEI_440}  \hyperref[html:y80]{terms_80}  \hyperref[html:y451]{front_gap_451}
                                  \hyperref[html:y130]{selbri_130}  \hyperref[html:y459]{SEhU_gap_459}
                        |  \hyperref[html:y440]{SEI_440}  \hyperref[html:y80]{terms_80}  \hyperref[html:y130]{selbri_130}  \hyperref[html:y459]{SEhU_gap_459}
                        ;

\phantomsection\label{html:y35}
vocative_35             :  \hyperref[html:y415]{DOI_415}  \hyperref[html:y130]{selbri_130}  \hyperref[html:y457]{DOhU_gap_457}
                        |  \hyperref[html:y415]{DOI_415}  \hyperref[html:y130]{selbri_130}
                                  \hyperref[html:y121]{relative_clauses_121}  \hyperref[html:y457]{DOhU_gap_457}
                        |  \hyperref[html:y415]{DOI_415}  \hyperref[html:y121]{relative_clauses_121}
                                  \hyperref[html:y130]{selbri_130}  \hyperref[html:y457]{DOhU_gap_457}
                        |  \hyperref[html:y415]{DOI_415}  \hyperref[html:y121]{relative_clauses_121}
                                  \hyperref[html:y130]{selbri_130}  \hyperref[html:y121]{relative_clauses_121}  \hyperref[html:y457]{DOhU_gap_457}
                        |  \hyperref[html:y415]{DOI_415}  \hyperref[html:y404]{cmene_404}  \hyperref[html:y457]{DOhU_gap_457}
                        |  \hyperref[html:y415]{DOI_415}  \hyperref[html:y404]{cmene_404}
                                  \hyperref[html:y121]{relative_clauses_121}  \hyperref[html:y457]{DOhU_gap_457}
                        |  \hyperref[html:y415]{DOI_415}  \hyperref[html:y121]{relative_clauses_121}  \hyperref[html:y404]{cmene_404}
                                 \hyperref[html:y457]{DOhU_gap_457}
                        |  \hyperref[html:y415]{DOI_415}  \hyperref[html:y121]{relative_clauses_121}  \hyperref[html:y404]{cmene_404}
                                  \hyperref[html:y121]{relative_clauses_121}  \hyperref[html:y457]{DOhU_gap_457}
                        |  \hyperref[html:y415]{DOI_415}  \hyperref[html:y90]{sumti_90}  \hyperref[html:y457]{DOhU_gap_457}
                        |  \hyperref[html:y415]{DOI_415}  \hyperref[html:y457]{DOhU_gap_457}
                        ;

\phantomsection\label{html:y36}
parenthetical_36        :  \hyperref[html:y606]{TO_606}  \hyperref[html:y0]{text_0}  \hyperref[html:y468]{TOI_gap_468}
                        ;

\phantomsection\label{html:y40}
sentence_40             :  \hyperref[html:y50]{bridi_tail_50}  /* bare observative or mo answer */
                        |  \hyperref[html:y80]{terms_80}  \hyperref[html:y451]{front_gap_451}  \hyperref[html:y50]{bridi_tail_50}
                        |  \hyperref[html:y80]{terms_80}  \hyperref[html:y50]{bridi_tail_50}
                        ;

\phantomsection\label{html:y41}
subsentence_41          :  \hyperref[html:y40]{sentence_40}
                        |  \hyperref[html:y30]{prenex_30}  \hyperref[html:y41]{subsentence_41}
                        ;

\phantomsection\label{html:y50}
bridi_tail_50           :  \hyperref[html:y51]{bridi_tail_A_51}
                        |  \hyperref[html:y51]{bridi_tail_A_51}  \hyperref[html:y814]{GIhEK_KE_814}  \hyperref[html:y50]{bridi_tail_50}
                                  \hyperref[html:y466]{KEhE_gap_466}  \hyperref[html:y71]{tail_terms_71}
                        ;

\phantomsection\label{html:y51}
bridi_tail_A_51         :  \hyperref[html:y52]{bridi_tail_B_52}
                        |  \hyperref[html:y51]{bridi_tail_A_51}  \hyperref[html:y818]{GIhEK_818}  \hyperref[html:y52]{bridi_tail_B_52}
                           \hyperref[html:y71]{tail_terms_71}
                        ;

\phantomsection\label{html:y52}
bridi_tail_B_52         :  \hyperref[html:y53]{bridi_tail_C_53}
                        |  \hyperref[html:y53]{bridi_tail_C_53}  \hyperref[html:y813]{GIhEK_BO_813}  \hyperref[html:y52]{bridi_tail_B_52}
                           \hyperref[html:y71]{tail_terms_71}
                        ;

\phantomsection\label{html:y53}
bridi_tail_C_53         :  \hyperref[html:y54]{gek_sentence_54}
                        |  \hyperref[html:y130]{selbri_130}  \hyperref[html:y71]{tail_terms_71}
                        ;

\phantomsection\label{html:y54}
gek_sentence_54         :  \hyperref[html:y807]{GEK_807}  \hyperref[html:y41]{subsentence_41}
                                \hyperref[html:y816]{GIK_816}  \hyperref[html:y41]{subsentence_41}  \hyperref[html:y71]{tail_terms_71}
                        |  \hyperref[html:y491]{tag_491}  \hyperref[html:y493]{KE_493}  \hyperref[html:y54]{gek_sentence_54}  \hyperref[html:y466]{KEhE_gap_466}
                        |  \hyperref[html:y445]{NA_445}  \hyperref[html:y54]{gek_sentence_54}
                        ;

\phantomsection\label{html:y71}
tail_terms_71           :  \hyperref[html:y80]{terms_80}  \hyperref[html:y456]{VAU_gap_456}
                        |  \hyperref[html:y456]{VAU_gap_456}
                        ;

\phantomsection\label{html:y80}
terms_80                :  \hyperref[html:y81]{terms_A_81}
                        |  \hyperref[html:y80]{terms_80}  \hyperref[html:y81]{terms_A_81}
                        ;

\phantomsection\label{html:y81}
terms_A_81              :  \hyperref[html:y82]{terms_B_82}
                        |  \hyperref[html:y81]{terms_A_81}  \hyperref[html:y494]{PEhE_494}  \hyperref[html:y422]{JOIK_JEK_422}  \hyperref[html:y82]{terms_B_82}
                        ;

\phantomsection\label{html:y82}
terms_B_82              :  \hyperref[html:y83]{term_83}
                        |  \hyperref[html:y82]{terms_B_82}  \hyperref[html:y495]{CEhE_495}  \hyperref[html:y83]{term_83}
                        ;

\phantomsection\label{html:y83}
term_83                 :  \hyperref[html:y90]{sumti_90}
                        |  \hyperref[html:y84]{modifier_84}
                        |  \hyperref[html:y85]{term_set_85}
                        |  \hyperref[html:y810]{NA_KU_810}
                        ;

\phantomsection\label{html:y84}
modifier_84             :  \hyperref[html:y490]{mod_head_490}  \hyperref[html:y450]{gap_450}
                        |  \hyperref[html:y490]{mod_head_490}  \hyperref[html:y90]{sumti_90}
                        ;

\phantomsection\label{html:y85}
term_set_85             :  \hyperref[html:y496]{NUhI_496}  \hyperref[html:y80]{terms_80}  \hyperref[html:y460]{NUhU_gap_460}
                        |  \hyperref[html:y496]{NUhI_496}  \hyperref[html:y807]{GEK_807}  \hyperref[html:y80]{terms_80}  \hyperref[html:y460]{NUhU_gap_460}
                                \hyperref[html:y816]{GIK_816}  \hyperref[html:y80]{terms_80}  \hyperref[html:y460]{NUhU_gap_460}
                        ;

\phantomsection\label{html:y90}
sumti_90                :  \hyperref[html:y91]{sumti_A_91}
                        |  \hyperref[html:y91]{sumti_A_91}  \hyperref[html:y497]{VUhO_497}  \hyperref[html:y121]{relative_clauses_121}
                        ;

\phantomsection\label{html:y91}
sumti_A_91              :  \hyperref[html:y92]{sumti_B_92}
                        |  \hyperref[html:y92]{sumti_B_92}  \hyperref[html:y804]{EK_KE_804}  \hyperref[html:y90]{sumti_90}  \hyperref[html:y466]{KEhE_gap_466}
                        |  \hyperref[html:y92]{sumti_B_92}  \hyperref[html:y823]{JOIK_KE_823}  \hyperref[html:y90]{sumti_90}  \hyperref[html:y466]{KEhE_gap_466}
                        ;

\phantomsection\label{html:y92}
sumti_B_92              :  \hyperref[html:y93]{sumti_C_93}
                        |  \hyperref[html:y92]{sumti_B_92}  \hyperref[html:y421]{JOIK_EK_421}  \hyperref[html:y93]{sumti_C_93}  
                        ;

\phantomsection\label{html:y93}
sumti_C_93              :  \hyperref[html:y94]{sumti_D_94}
                        |  \hyperref[html:y94]{sumti_D_94}  \hyperref[html:y803]{EK_BO_803}  \hyperref[html:y93]{sumti_C_93}
                        |  \hyperref[html:y94]{sumti_D_94}  \hyperref[html:y822]{JOIK_BO_822}  \hyperref[html:y93]{sumti_C_93}
                        ;

\phantomsection\label{html:y94}
sumti_D_94              :  \hyperref[html:y95]{sumti_E_95}
                        |  \hyperref[html:y807]{GEK_807}  \hyperref[html:y90]{sumti_90}  \hyperref[html:y816]{GIK_816}  \hyperref[html:y94]{sumti_D_94}
                        ;

\phantomsection\label{html:y95}
sumti_E_95              :  \hyperref[html:y96]{sumti_F_96}
                        |  \hyperref[html:y96]{sumti_F_96}  \hyperref[html:y121]{relative_clauses_121}
                           /* indefinite sumti */
                        |  \hyperref[html:y300]{quantifier_300}  \hyperref[html:y130]{selbri_130}  \hyperref[html:y450]{gap_450}
                        |  \hyperref[html:y300]{quantifier_300}  \hyperref[html:y130]{selbri_130}
                                \hyperref[html:y450]{gap_450}  \hyperref[html:y121]{relative_clauses_121}
                        ;

\phantomsection\label{html:y96}
sumti_F_96              :  \hyperref[html:y97]{sumti_G_97}
                           /* outer-quantified sumti */
                        |  \hyperref[html:y300]{quantifier_300}  \hyperref[html:y97]{sumti_G_97}
                        ;

\phantomsection\label{html:y97}
sumti_G_97              :  \hyperref[html:y483]{qualifier_483}  \hyperref[html:y90]{sumti_90}  \hyperref[html:y463]{LUhU_gap_463}
                        |  \hyperref[html:y483]{qualifier_483}  \hyperref[html:y121]{relative_clauses_121}
                                \hyperref[html:y90]{sumti_90}  \hyperref[html:y463]{LUhU_gap_463}
                           /*sumti grouping, set/mass/individual conversion */
                           /*also sumti scalar negation */
                        |  \hyperref[html:y400]{anaphora_400}
                        |  \hyperref[html:y499]{LA_499}  \hyperref[html:y404]{cmene_404}
                        |  \hyperref[html:y499]{LA_499}  \hyperref[html:y121]{relative_clauses_121}  \hyperref[html:y404]{cmene_404}
                        |  \hyperref[html:y489]{LI_489}  \hyperref[html:y310]{MEX_310}  \hyperref[html:y472]{LOhO_gap_472}
                        |  \hyperref[html:y110]{description_110}
                        |  \hyperref[html:y432]{quote_arg_432}
                        ;

\phantomsection\label{html:y110}
description_110         :  \hyperref[html:y499]{LA_499}  \hyperref[html:y111]{sumti_tail_111}  \hyperref[html:y450]{gap_450}
                        |  \hyperref[html:y488]{LE_488}  \hyperref[html:y111]{sumti_tail_111}  \hyperref[html:y450]{gap_450}
                        ;

\phantomsection\label{html:y111}
sumti_tail_111          :  \hyperref[html:y112]{sumti_tail_A_112}
                           /* inner-quantified sumti relative clause */
                        |  \hyperref[html:y121]{relative_clauses_121}  \hyperref[html:y112]{sumti_tail_A_112}
                           /* pseudo-possessive
                              (an abbreviated inner restriction);
                              note that sumti cannot be quantified */
                        |  \hyperref[html:y97]{sumti_G_97}  \hyperref[html:y112]{sumti_tail_A_112}
                           /* pseudo-possessive with outer restriction */
                        |  \hyperref[html:y97]{sumti_G_97}  \hyperref[html:y121]{relative_clauses_121}  \hyperref[html:y112]{sumti_tail_A_112}
                        ;

\phantomsection\label{html:y112}
sumti_tail_A_112        :  \hyperref[html:y130]{selbri_130}
                        |  \hyperref[html:y130]{selbri_130}  \hyperref[html:y121]{relative_clauses_121}
                           /* explicit inner quantifier */
                        |  \hyperref[html:y300]{quantifier_300}  \hyperref[html:y130]{selbri_130}
                           /* quantifier both internal to a description,
                              and external to a sumti thereby made specific */
                        |  \hyperref[html:y300]{quantifier_300}  \hyperref[html:y130]{selbri_130}  \hyperref[html:y121]{relative_clauses_121}
                        |  \hyperref[html:y300]{quantifier_300}  \hyperref[html:y90]{sumti_90}
                        ;

\phantomsection\label{html:y121}
relative_clauses_121    :  \hyperref[html:y122]{relative_clause_122}
                        |  \hyperref[html:y121]{relative_clauses_121}  \hyperref[html:y487]{ZIhE_487}  \hyperref[html:y122]{relative_clause_122}
                        ;

\phantomsection\label{html:y122}
relative_clause_122     :  \hyperref[html:y485]{GOI_485}  \hyperref[html:y83]{term_83}  \hyperref[html:y464]{GEhU_gap_464}
                        |  \hyperref[html:y484]{NOI_484}  \hyperref[html:y41]{subsentence_41}  \hyperref[html:y469]{KUhO_gap_469}
                        ;

\phantomsection\label{html:y130}
selbri_130              :  \hyperref[html:y491]{tag_491}  \hyperref[html:y131]{selbri_A_131}
                        |  \hyperref[html:y131]{selbri_A_131}
                        ;

\phantomsection\label{html:y131}
selbri_A_131            :  \hyperref[html:y132]{selbri_B_132}
                        |  \hyperref[html:y445]{NA_445}  \hyperref[html:y130]{selbri_130}
                        ;

\phantomsection\label{html:y132}
selbri_B_132            :  \hyperref[html:y133]{selbri_C_133}
                        |  \hyperref[html:y133]{selbri_C_133}  \hyperref[html:y443]{CO_443}  \hyperref[html:y132]{selbri_B_132}
                        ;

\phantomsection\label{html:y133}
selbri_C_133            :  \hyperref[html:y134]{selbri_D_134}
                        |  \hyperref[html:y133]{selbri_C_133}  \hyperref[html:y134]{selbri_D_134}
                        ;

\phantomsection\label{html:y134}
selbri_D_134            :  \hyperref[html:y135]{selbri_E_135}
                        |  \hyperref[html:y134]{selbri_D_134}  \hyperref[html:y422]{JOIK_JEK_422}  \hyperref[html:y135]{selbri_E_135}
                        |  \hyperref[html:y134]{selbri_D_134}  \hyperref[html:y823]{JOIK_KE_823}  \hyperref[html:y133]{selbri_C_133}
                                \hyperref[html:y466]{KEhE_gap_466}
                        ;

\phantomsection\label{html:y135}
selbri_E_135            :  \hyperref[html:y136]{selbri_F_136}
                        |  \hyperref[html:y136]{selbri_F_136}  \hyperref[html:y821]{JEK_BO_821}  \hyperref[html:y135]{selbri_E_135}
                        |  \hyperref[html:y136]{selbri_F_136}  \hyperref[html:y822]{JOIK_BO_822}  \hyperref[html:y135]{selbri_E_135}
                        ;

\phantomsection\label{html:y136}
selbri_F_136            :  \hyperref[html:y150]{tanru_unit_150}
                        |  \hyperref[html:y150]{tanru_unit_150}  \hyperref[html:y479]{BO_479}  \hyperref[html:y136]{selbri_F_136}
                        |  \hyperref[html:y137]{GUhEK_selbri_137}
                        |  \hyperref[html:y482]{NAhE_482}  \hyperref[html:y137]{GUhEK_selbri_137}
                        ;

\phantomsection\label{html:y137}
GUhEK_selbri_137        :  \hyperref[html:y808]{GUhEK_808}  \hyperref[html:y130]{selbri_130}  \hyperref[html:y816]{GIK_816}  \hyperref[html:y136]{selbri_F_136}
                        ;

\phantomsection\label{html:y150}
tanru_unit_150          :  \hyperref[html:y151]{tanru_unit_A_151}
                        |  \hyperref[html:y150]{tanru_unit_150}  \hyperref[html:y444]{CEI_444}  \hyperref[html:y151]{tanru_unit_A_151}
                        ;

\phantomsection\label{html:y151}
tanru_unit_A_151        :  \hyperref[html:y152]{tanru_unit_B_152}
                        |  \hyperref[html:y152]{tanru_unit_B_152}  \hyperref[html:y160]{linkargs_160}
                        ;

\phantomsection\label{html:y152}
tanru_unit_B_152        :  \hyperref[html:y407]{bridi_valsi_407}
                        |  \hyperref[html:y493]{KE_493}  \hyperref[html:y133]{selbri_C_133}  \hyperref[html:y466]{KEhE_gap_466}
                        |  \hyperref[html:y480]{SE_480}  \hyperref[html:y152]{tanru_unit_B_152}
                        |  \hyperref[html:y478]{JAI_478}  \hyperref[html:y491]{tag_491}  \hyperref[html:y152]{tanru_unit_B_152}
                        |  \hyperref[html:y478]{JAI_478}  \hyperref[html:y152]{tanru_unit_B_152}
                        |  \hyperref[html:y477]{ME_477}  \hyperref[html:y90]{sumti_90}  \hyperref[html:y465]{MEhU_gap_465}
                        |  \hyperref[html:y477]{ME_477}  \hyperref[html:y90]{sumti_90}  \hyperref[html:y465]{MEhU_gap_465}  \hyperref[html:y476]{MOI_476}
                        |  \hyperref[html:y475]{NUhA_475}  \hyperref[html:y374]{MEX_operator_374}
                        |  \hyperref[html:y482]{NAhE_482}  \hyperref[html:y152]{tanru_unit_B_152}
                        |  \hyperref[html:y425]{NU_425}  \hyperref[html:y41]{subsentence_41}  \hyperref[html:y453]{KEI_gap_453}
                        ;

\phantomsection\label{html:y160}
linkargs_160            :  \hyperref[html:y446]{BE_446}  \hyperref[html:y83]{term_83}  \hyperref[html:y467]{BEhO_gap_467}
                        |  \hyperref[html:y446]{BE_446}  \hyperref[html:y83]{term_83}  \hyperref[html:y161]{links_161}  \hyperref[html:y467]{BEhO_gap_467}
                        ;

\phantomsection\label{html:y161}
links_161               :  \hyperref[html:y442]{BEI_442}  \hyperref[html:y83]{term_83}
                        |  \hyperref[html:y442]{BEI_442}  \hyperref[html:y83]{term_83}  \hyperref[html:y161]{links_161}
                        ;

/*  Main entry point for MEX; everything but a number must be in parens.   */

\phantomsection\label{html:y300}
quantifier_300          :  \hyperref[html:y812]{number_812}  \hyperref[html:y461]{BOI_gap_461}
                        |  \hyperref[html:y470]{left_bracket_470}  \hyperref[html:y310]{MEX_310}  \hyperref[html:y471]{right_bracket_gap_471}
                        ;

/*  Entry point for MEX used after LI; no parens needed, but LI now has an
    elidable terminator. (This allows us to express the difference between
    \q{the expression a + b} and \q{the expression (a + b)}_)   */

/*  This rule supports left-grouping infix expressions and reverse Polish
    expressions. To handle infix monadic, use a null operand; to handle
    infix with more than two operands (whatever that means) use an extra
    operator or an array operand.   */

\phantomsection\label{html:y310}
MEX_310                 :  \hyperref[html:y311]{MEX_A_311}
                        |  \hyperref[html:y310]{MEX_310}  \hyperref[html:y370]{operator_370}  \hyperref[html:y311]{MEX_A_311}
                        |  \hyperref[html:y441]{FUhA_441}  \hyperref[html:y330]{rp_expression_330}
                        ;

/*  Support for right-grouping (short scope) infix expressions with BIhE.  */

\phantomsection\label{html:y311}
MEX_A_311               :  \hyperref[html:y312]{MEX_B_312}
                        |  \hyperref[html:y312]{MEX_B_312}  \hyperref[html:y439]{BIhE_439}  \hyperref[html:y370]{operator_370}  \hyperref[html:y311]{MEX_A_311}
                        ;

/*  Support for forethought (Polish) expressions. These begin with a
    forethought flag, then the operator and then the argument(s).  */

\phantomsection\label{html:y312}
MEX_B_312               :  \hyperref[html:y381]{operand_381}
                        |  \hyperref[html:y370]{operator_370}  \hyperref[html:y313]{MEX_C_313}  \hyperref[html:y452]{MEX_gap_452}
                        |  \hyperref[html:y438]{PEhO_438}  \hyperref[html:y370]{operator_370}  \hyperref[html:y313]{MEX_C_313}  \hyperref[html:y452]{MEX_gap_452}
                        ;

\phantomsection\label{html:y313}
MEX_C_313               :  \hyperref[html:y312]{MEX_B_312}
                        |  \hyperref[html:y313]{MEX_C_313}  \hyperref[html:y312]{MEX_B_312}
                        ;

/*  Reverse Polish expressions always have exactly two operands.
    To handle one operand, use a null operand;
    to handle more than two operands, use a null operator.  */

\phantomsection\label{html:y330}
rp_expression_330       :  \hyperref[html:y332]{rp_operand_332}  \hyperref[html:y332]{rp_operand_332}  \hyperref[html:y370]{operator_370}
                        ;

\phantomsection\label{html:y332}
rp_operand_332          :  \hyperref[html:y381]{operand_381}
                        |  \hyperref[html:y330]{rp_expression_330}
                        ;

/*  Operators may be joined by logical connectives. */

\phantomsection\label{html:y370}
operator_370            :  \hyperref[html:y371]{operator_A_371}
                        |  \hyperref[html:y370]{operator_370}  \hyperref[html:y422]{JOIK_JEK_422}  \hyperref[html:y371]{operator_A_371}
                        |  \hyperref[html:y370]{operator_370}  \hyperref[html:y823]{JOIK_KE_823}  \hyperref[html:y370]{operator_370}
                                \hyperref[html:y466]{KEhE_gap_466}
                        ;

\phantomsection\label{html:y371}
operator_A_371          :  \hyperref[html:y372]{operator_B_372}
                        |  \hyperref[html:y808]{GUhEK_808}  \hyperref[html:y371]{operator_A_371}  \hyperref[html:y816]{GIK_816}  \hyperref[html:y372]{operator_B_372}
                        |  \hyperref[html:y372]{operator_B_372}  \hyperref[html:y822]{JOIK_BO_822}  \hyperref[html:y371]{operator_A_371}
                        |  \hyperref[html:y372]{operator_B_372}  \hyperref[html:y821]{JEK_BO_821}  \hyperref[html:y371]{operator_A_371}
                        ;

\phantomsection\label{html:y372}
operator_B_372          :  \hyperref[html:y374]{MEX_operator_374}
                        |  \hyperref[html:y493]{KE_493}  \hyperref[html:y370]{operator_370}  \hyperref[html:y466]{KEhE_gap_466}
                        ;

\phantomsection\label{html:y374}
MEX_operator_374        :  \hyperref[html:y679]{VUhU_679}
                        |  \hyperref[html:y679]{VUhU_679}  \hyperref[html:y32]{free_modifier_32}
                        |  \hyperref[html:y480]{SE_480}  \hyperref[html:y374]{MEX_operator_374}
   /* changes argument order */
                        |  \hyperref[html:y482]{NAhE_482}  \hyperref[html:y374]{MEX_operator_374}
   /* scalar negation */
                        |  \hyperref[html:y430]{MAhO_430}  \hyperref[html:y310]{MEX_310}  \hyperref[html:y473]{TEhU_gap_473}
                        |  \hyperref[html:y429]{NAhU_429}  \hyperref[html:y130]{selbri_130}  \hyperref[html:y473]{TEhU_gap_473}
                        ;

\phantomsection\label{html:y381}
operand_381             :  \hyperref[html:y382]{operand_A_382}
                        |  \hyperref[html:y382]{operand_A_382}  \hyperref[html:y804]{EK_KE_804}  \hyperref[html:y381]{operand_381}  \hyperref[html:y466]{KEhE_gap_466}
                        |  \hyperref[html:y382]{operand_A_382}  \hyperref[html:y823]{JOIK_KE_823}  \hyperref[html:y381]{operand_381}  \hyperref[html:y466]{KEhE_gap_466}
                        ;

\phantomsection\label{html:y382}
operand_A_382           :  \hyperref[html:y383]{operand_B_383}
                        |  \hyperref[html:y382]{operand_A_382}  \hyperref[html:y421]{JOIK_EK_421}  \hyperref[html:y383]{operand_B_383}
                        ;

\phantomsection\label{html:y383}
operand_B_383           :  \hyperref[html:y385]{operand_C_385}
                        |  \hyperref[html:y385]{operand_C_385}  \hyperref[html:y803]{EK_BO_803}  \hyperref[html:y383]{operand_B_383}
                        |  \hyperref[html:y385]{operand_C_385}  \hyperref[html:y822]{JOIK_BO_822}  \hyperref[html:y383]{operand_B_383}
                        ;

\phantomsection\label{html:y385}
operand_C_385           :  \hyperref[html:y300]{quantifier_300}
                        |  \hyperref[html:y817]{lerfu_string_817}  \hyperref[html:y461]{BOI_gap_461}
   /* lerfu string as operand - classic math variable */
                        |  \hyperref[html:y428]{NIhE_428}  \hyperref[html:y130]{selbri_130}  \hyperref[html:y473]{TEhU_gap_473}
   /* quantifies a bridi - inverse of -MOI */
                        |  \hyperref[html:y427]{MOhE_427}  \hyperref[html:y90]{sumti_90}  \hyperref[html:y473]{TEhU_gap_473}
   /* quantifies a sumti - inverse of LI */
                        |  \hyperref[html:y431]{JOhI_431}  \hyperref[html:y313]{MEX_C_313}  \hyperref[html:y473]{TEhU_gap_473}
                        |  \hyperref[html:y807]{GEK_807}  \hyperref[html:y381]{operand_381}  \hyperref[html:y816]{GIK_816}  \hyperref[html:y385]{operand_C_385}
                        |  \hyperref[html:y483]{qualifier_483}  \hyperref[html:y381]{operand_381}  \hyperref[html:y463]{LUhU_gap_463}
                        ;

/* _400 series constructs are mostly specific strings, some of which may
   also be used by the lexer; the lexer should not use any reference to
   terminals numbered less than _400, as they have grammars composed on
   non-deterministic strings of selma'o.  Some above _400 also are this
   way, so care should be taken; this is especially true for those that
   reference \hyperref[html:y32]{free_modifier_32}.  */

\phantomsection\label{html:y400}
anaphora_400            :  \hyperref[html:y555]{KOhA_555}
                        |  \hyperref[html:y555]{KOhA_555}  \hyperref[html:y32]{free_modifier_32}
                        |  \hyperref[html:y817]{lerfu_string_817}  \hyperref[html:y461]{BOI_gap_461}
                        ;

\phantomsection\label{html:y404}
cmene_404               :  \hyperref[html:y405]{cmene_A_405}
                        |  \hyperref[html:y405]{cmene_A_405}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y405}
cmene_A_405             :  \hyperref[html:y518]{CMENE_518}  /* pause */
                        |  \hyperref[html:y405]{cmene_A_405}  \hyperref[html:y518]{CMENE_518}  /* pause*/
/* multiple CMENE are identified morphologically (by the lexer) -- separated by
   consonant \&pause */
                        ;

\phantomsection\label{html:y407}
bridi_valsi_407         :  \hyperref[html:y408]{bridi_valsi_A_408}
                        |  \hyperref[html:y408]{bridi_valsi_A_408}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y408}
bridi_valsi_A_408       :  \hyperref[html:y509]{BRIVLA_509}
                        |  \hyperref[html:y824]{PA_MOI_824}
                        |  \hyperref[html:y543]{GOhA_543}
                        |  \hyperref[html:y543]{GOhA_543}  \hyperref[html:y593]{RAhO_593}
                        ;

\phantomsection\label{html:y410}
para_mark_410           :  \hyperref[html:y584]{NIhO_584}
                        |  \hyperref[html:y584]{NIhO_584}  \hyperref[html:y32]{free_modifier_32}
                        |  \hyperref[html:y584]{NIhO_584}  \hyperref[html:y410]{para_mark_410}
                        ;

\phantomsection\label{html:y411}
indicators_411          :  \hyperref[html:y412]{indicators_A_412}
                        |  \hyperref[html:y535]{FUhE_535}  \hyperref[html:y412]{indicators_A_412}
                        ;

\phantomsection\label{html:y412}
indicators_A_412        :  \hyperref[html:y413]{indicator_413}
                        |  \hyperref[html:y412]{indicators_A_412}  \hyperref[html:y413]{indicator_413}
                        ;

\phantomsection\label{html:y413}
indicator_413           :  \hyperref[html:y612]{UI_612}
                        |  \hyperref[html:y515]{CAI_515}
                        |  \hyperref[html:y612]{UI_612}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y515]{CAI_515}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y619]{Y_619}
                        |  \hyperref[html:y524]{DAhO_524}
                        |  \hyperref[html:y536]{FUhO_536}
                        ;

\phantomsection\label{html:y415}
DOI_415                 :  \hyperref[html:y525]{DOI_525}
                        |  \hyperref[html:y416]{COI_416}
                        |  \hyperref[html:y416]{COI_416}  \hyperref[html:y525]{DOI_525}
                        ;

\phantomsection\label{html:y416}
COI_416                 :  \hyperref[html:y417]{COI_A_417}
                        |  \hyperref[html:y416]{COI_416}  \hyperref[html:y417]{COI_A_417}
                        ;

\phantomsection\label{html:y417}
COI_A_417               :  \hyperref[html:y520]{COI_520}
                        |  \hyperref[html:y520]{COI_520}  \hyperref[html:y581]{NAI_581}
                        ;

\phantomsection\label{html:y421}
JOIK_EK_421             :  \hyperref[html:y802]{EK_802}
                        |  \hyperref[html:y806]{JOIK_806}
                        |  \hyperref[html:y806]{JOIK_806}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y422}
JOIK_JEK_422            :  \hyperref[html:y806]{JOIK_806}
                        |  \hyperref[html:y806]{JOIK_806}  \hyperref[html:y32]{free_modifier_32}
                        |  \hyperref[html:y805]{JEK_805}
                        |  \hyperref[html:y805]{JEK_805}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y424}
XI_424                  :  \hyperref[html:y618]{XI_618}
                        |  \hyperref[html:y618]{XI_618}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y425}
NU_425                  :  \hyperref[html:y426]{NU_A_426}
                        |  \hyperref[html:y425]{NU_425}  \hyperref[html:y422]{JOIK_JEK_422}  \hyperref[html:y426]{NU_A_426}
                        ;

\phantomsection\label{html:y426}
NU_A_426                :  \hyperref[html:y586]{NU_586}
                        |  \hyperref[html:y586]{NU_586}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y586]{NU_586}  \hyperref[html:y32]{free_modifier_32}
                        |  \hyperref[html:y586]{NU_586}  \hyperref[html:y581]{NAI_581}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y427}
MOhE_427                :  \hyperref[html:y664]{MOhE_664}
                        |  \hyperref[html:y664]{MOhE_664}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y428}
NIhE_428                :  \hyperref[html:y666]{NIhE_666}
                        |  \hyperref[html:y666]{NIhE_666}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y429}
NAhU_429                :  \hyperref[html:y665]{NAhU_665}
                        |  \hyperref[html:y665]{NAhU_665}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y430}
MAhO_430                :  \hyperref[html:y662]{MAhO_662}
                        |  \hyperref[html:y662]{MAhO_662}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y431}
JOhI_431                :  \hyperref[html:y657]{JOhI_657}
                        |  \hyperref[html:y657]{JOhI_657}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y432}
quote_arg_432           :  \hyperref[html:y433]{quote_arg_A_433}
                        |  \hyperref[html:y433]{quote_arg_A_433}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y433}
quote_arg_A_433         :  \hyperref[html:y434]{ZOI_quote_434}
                        |  \hyperref[html:y435]{ZO_quote_435}
                        |  \hyperref[html:y436]{LOhU_quote_436}
                        |  \hyperref[html:y571]{LU_571}  \hyperref[html:y0]{text_0}  \hyperref[html:y448]{LIhU_gap_448}
                        ;

/* The quoted material in the following three terminals must be identified by
   the lexer, but no additional lexer processing is needed. */

\phantomsection\label{html:y434}
ZOI_quote_434           :  \hyperref[html:y627]{ZOI_627}  \hyperref[html:y698]{any_word_698}
                                  /*pause*/  \hyperref[html:y699]{anything_699}  /*pause*/  \hyperref[html:y698]{any_word_698}
                        ;

/* 'pause' is morphemic, represented by '.' The lexer assembles \hyperref[html:y699]{anything_699} */

\phantomsection\label{html:y435}
ZO_quote_435            :  \hyperref[html:y626]{ZO_626}  \hyperref[html:y698]{any_word_698}
                        ;

/* 'word' may not be a compound; but it can be any valid Lojban selma'o value,
   including ZO, ZOI, SI, SA, SU. The preparser will not lex the word per its
   normal selma'o. */

\phantomsection\label{html:y436}
LOhU_quote_436          :  \hyperref[html:y569]{LOhU_569}  \hyperref[html:y697]{any_words_697}  \hyperref[html:y565]{LEhU_565}
                        ;

/* 'words' may be any Lojban words, with no claim of grammaticality; the
   preparser will not lex the individual words per their normal selma'o;
   used to quote ungrammatical Lojban, equivalent to the * or ? writing
   convention for such text.  */

/* The preparser needs one bit of sophistication for this rule.  A
   quoted string should be able to contain other quoted strings - this is
   only a problem for a LOhU quote itself, since the LEhU clossing this
   quote would otherwise close the outer quotes, which is incorrect.  For
   this purpose, we will cheat on the use of ZO in such a quote (since this
   is ungrammatical text, it is a sign ignored by the parser).  Use ZO to
   mark any nested quotation LOhU.  The preparser then will absorb it by
   the ZO rule, before testing for LOhU.  This is obviously not the
   standard usage for ZO, which would otherwise cause the result to be a
   sumti.  But, since the result will be part of an unparsed string anyway,
   it doesn't matter.  */

/* It may be seen that any of the ZO/ZOI/LOhU trio of quotation markers
   may contain the powerful metalinguistic erasers.  Since these quotations
   are not parsed internally, these operators are ignored within the quote.
   To erase a ZO, then, two SI's are needed after giving a quoted word of
   any type.  ZOI takes four SI's, with the ENTIRE BODY OF THE QUOTE
   treated as a single 'word' since it is one selma'o.  Thus one for the
   quote body, two for the single word delimiters, and one for the ZOI.  In
   LOhU, the entire body is treated as a single word, so three SI's can
   erase it.  */

/* All rule terminator names with 'gap' in them are potentially
   elidable, where such elision does not cause an ambiguity.  This is
   implemented through use of the YACC 'error' token, which effectively
   recovers from an elision.  */

\phantomsection\label{html:y437}
FIhO_437                :  \hyperref[html:y532]{FIhO_532}
                        |  \hyperref[html:y532]{FIhO_532}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y438}
PEhO_438                :  \hyperref[html:y673]{PEhO_673}
                        |  \hyperref[html:y673]{PEhO_673}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y439}
BIhE_439                :  \hyperref[html:y650]{BIhE_650}
                        |  \hyperref[html:y650]{BIhE_650}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y440}
SEI_440                 :  \hyperref[html:y597]{SEI_597}
                        |  \hyperref[html:y597]{SEI_597}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y441}
FUhA_441                :  \hyperref[html:y655]{FUhA_655}
                        |  \hyperref[html:y655]{FUhA_655}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y442}
BEI_442                 :  \hyperref[html:y505]{BEI_505}
                        |  \hyperref[html:y505]{BEI_505}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y443}
CO_443                  :  \hyperref[html:y519]{CO_519}
                        |  \hyperref[html:y519]{CO_519}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y444}
CEI_444                 :  \hyperref[html:y516]{CEI_516}
                        |  \hyperref[html:y516]{CEI_516}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y445}
NA_445                  :  \hyperref[html:y578]{NA_578}
                        |  \hyperref[html:y578]{NA_578}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y446}
BE_446                  :  \hyperref[html:y504]{BE_504}
                        |  \hyperref[html:y504]{BE_504}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y447}
TUhE_447                :  \hyperref[html:y610]{TUhE_610}
                        |  \hyperref[html:y610]{TUhE_610}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y448}
LIhU_gap_448            :  \hyperref[html:y567]{LIhU_567}
                        |  error
                        ;

\phantomsection\label{html:y450}
gap_450                 :  \hyperref[html:y556]{KU_556}
                        |  \hyperref[html:y556]{KU_556}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y451}
front_gap_451           :  \hyperref[html:y521]{CU_521}
                        |  \hyperref[html:y521]{CU_521}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y452}
MEX_gap_452             :  \hyperref[html:y658]{KUhE_658}
                        |  \hyperref[html:y658]{KUhE_658}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y453}
KEI_gap_453             :  \hyperref[html:y552]{KEI_552}
                        |  \hyperref[html:y552]{KEI_552}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y454}
TUhU_gap_454            :  \hyperref[html:y611]{TUhU_611}
                        |  \hyperref[html:y611]{TUhU_611}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y456}
VAU_gap_456             :  \hyperref[html:y614]{VAU_614}
                        |  \hyperref[html:y614]{VAU_614}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

/* redundant to attach a free modifier on the following */

\phantomsection\label{html:y457}
DOhU_gap_457            :  \hyperref[html:y526]{DOhU_526}
                        |  error
                        ;

\phantomsection\label{html:y458}
FEhU_gap_458            :  \hyperref[html:y531]{FEhU_531}
                        |  \hyperref[html:y531]{FEhU_531}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y459}
SEhU_gap_459            :  \hyperref[html:y598]{SEhU_598}
                        |  error
/* a free modifier on a discursive should be somewhere within the discursive.
   See \hyperref[html:y440]{SEI_440} */
                        ;

\phantomsection\label{html:y460}
NUhU_gap_460            :  \hyperref[html:y588]{NUhU_588}
                        |  \hyperref[html:y588]{NUhU_588}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y461}
BOI_gap_461             :  \hyperref[html:y651]{BOI_651}
                        |  \hyperref[html:y651]{BOI_651}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y462}
sub_gap_462             :  \hyperref[html:y651]{BOI_651}
                        |  error
                        ;

\phantomsection\label{html:y463}
LUhU_gap_463            :  \hyperref[html:y573]{LUhU_573}
                        |  \hyperref[html:y573]{LUhU_573}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y464}
GEhU_gap_464            :  \hyperref[html:y538]{GEhU_538}
                        |  \hyperref[html:y538]{GEhU_538}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y465}
MEhU_gap_465            :  \hyperref[html:y575]{MEhU_575}
                        |  \hyperref[html:y575]{MEhU_575}   \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y466}
KEhE_gap_466            :  \hyperref[html:y550]{KEhE_550}
                        |  \hyperref[html:y550]{KEhE_550}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y467}
BEhO_gap_467            :  \hyperref[html:y506]{BEhO_506}
                        |  \hyperref[html:y506]{BEhO_506}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y468}
TOI_gap_468             :  \hyperref[html:y607]{TOI_607}
                        |  error
                        ;

\phantomsection\label{html:y469}
KUhO_gap_469            :  \hyperref[html:y557]{KUhO_557}
                        |  \hyperref[html:y557]{KUhO_557}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y470}
left_bracket_470        :  \hyperref[html:y677]{VEI_677}
                        |  \hyperref[html:y677]{VEI_677}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y471}
right_bracket_gap_471   :  \hyperref[html:y678]{VEhO_678}
                        |  \hyperref[html:y678]{VEhO_678}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y472}
LOhO_gap_472            :  \hyperref[html:y568]{LOhO_568}
                        |  \hyperref[html:y568]{LOhO_568}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y473}
TEhU_gap_473            :  \hyperref[html:y675]{TEhU_675}
                        |  \hyperref[html:y675]{TEhU_675}  \hyperref[html:y32]{free_modifier_32}
                        |  error
                        ;

\phantomsection\label{html:y474}
right_br_no_free_474    :  \hyperref[html:y678]{VEhO_678}
                        |  error
                        ;

\phantomsection\label{html:y475}
NUhA_475                :  \hyperref[html:y667]{NUhA_667}
                        |  \hyperref[html:y667]{NUhA_667}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y476}
MOI_476                 :  \hyperref[html:y663]{MOI_663}
                        |  \hyperref[html:y663]{MOI_663}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y477}
ME_477                  :  \hyperref[html:y574]{ME_574}
                        |  \hyperref[html:y574]{ME_574}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y478}
JAI_478                 :  \hyperref[html:y547]{JAI_547}
                        |  \hyperref[html:y547]{JAI_547}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y479}
BO_479                  :  \hyperref[html:y508]{BO_508}
                        |  \hyperref[html:y508]{BO_508}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y480}
SE_480                  :  \hyperref[html:y596]{SE_596}
                        |  \hyperref[html:y596]{SE_596}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y481}
FA_481                  :  \hyperref[html:y527]{FA_527}
                        |  \hyperref[html:y527]{FA_527}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y482}
NAhE_482                :  \hyperref[html:y583]{NAhE_583}
                        |  \hyperref[html:y583]{NAhE_583}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y483}
qualifier_483           :  \hyperref[html:y561]{LAhE_561}
                        |  \hyperref[html:y561]{LAhE_561}  \hyperref[html:y32]{free_modifier_32}
                        |  \hyperref[html:y809]{NAhE_BO_809}
                        ;

\phantomsection\label{html:y484}
NOI_484                 :  \hyperref[html:y585]{NOI_585}
                        |  \hyperref[html:y585]{NOI_585}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y485}
GOI_485                 :  \hyperref[html:y542]{GOI_542}
                        |  \hyperref[html:y542]{GOI_542}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y486}
subscript_486           :  \hyperref[html:y424]{XI_424}  \hyperref[html:y812]{number_812}  \hyperref[html:y462]{sub_gap_462}
                        |  \hyperref[html:y424]{XI_424}  \hyperref[html:y470]{left_bracket_470}  \hyperref[html:y310]{MEX_310}
                                \hyperref[html:y474]{right_br_no_free_474}
                        |  \hyperref[html:y424]{XI_424}  \hyperref[html:y817]{lerfu_string_817}  \hyperref[html:y462]{sub_gap_462}
                        ;

\phantomsection\label{html:y487}
ZIhE_487                :  \hyperref[html:y625]{ZIhE_625}
                        |  \hyperref[html:y625]{ZIhE_625}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y488}
LE_488                  :  \hyperref[html:y562]{LE_562}
                        |  \hyperref[html:y562]{LE_562}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y489}
LI_489                  :  \hyperref[html:y566]{LI_566}
                        |  \hyperref[html:y566]{LI_566}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y490}
mod_head_490            :  \hyperref[html:y491]{tag_491}
                        |  \hyperref[html:y481]{FA_481}
                        ;

\phantomsection\label{html:y491}
tag_491                 :  \hyperref[html:y815]{tense_modal_815}
                        |  \hyperref[html:y491]{tag_491}  \hyperref[html:y422]{JOIK_JEK_422}  \hyperref[html:y815]{tense_modal_815}
                        ;

\phantomsection\label{html:y492}
ZOhU_492                :  \hyperref[html:y628]{ZOhU_628}
                        |  \hyperref[html:y628]{ZOhU_628}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y493}
KE_493                  :  \hyperref[html:y551]{KE_551}
                        |  \hyperref[html:y551]{KE_551}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y494}
PEhE_494                :  \hyperref[html:y591]{PEhE_591}
                        |  \hyperref[html:y591]{PEhE_591}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y495}
CEhE_495                :  \hyperref[html:y517]{CEhE_517}
                        |  \hyperref[html:y517]{CEhE_517}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y496}
NUhI_496                :  \hyperref[html:y587]{NUhI_587}
                        |  \hyperref[html:y587]{NUhI_587}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y497}
VUhO_497                :  \hyperref[html:y617]{VUhO_617}
                        |  \hyperref[html:y617]{VUhO_617}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y498}
SOI_498                 :  \hyperref[html:y602]{SOI_602}
                        |  \hyperref[html:y602]{SOI_602}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y499}
LA_499                  :  \hyperref[html:y558]{LA_558}
                        |  \hyperref[html:y558]{LA_558}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y801}
utterance_ordinal_801   :  \hyperref[html:y905]{lexer_A_905}
                        ;

\phantomsection\label{html:y802}
EK_802                  :  \hyperref[html:y910]{lexer_B_910}
                        |  \hyperref[html:y910]{lexer_B_910}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y803}
EK_BO_803               :  \hyperref[html:y915]{lexer_C_915}
                        |  \hyperref[html:y915]{lexer_C_915}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y804}
EK_KE_804               :  \hyperref[html:y916]{lexer_D_916}
                        |  \hyperref[html:y916]{lexer_D_916}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y805}
JEK_805                 :  \hyperref[html:y925]{lexer_E_925}
                        ;

\phantomsection\label{html:y806}
JOIK_806                :  \hyperref[html:y930]{lexer_F_930}
                        ;

\phantomsection\label{html:y807}
GEK_807                 :  \hyperref[html:y935]{lexer_G_935}
                        |  \hyperref[html:y935]{lexer_G_935}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y808}
GUhEK_808               :  \hyperref[html:y940]{lexer_H_940}
                        |  \hyperref[html:y940]{lexer_H_940}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y809}
NAhE_BO_809             :  \hyperref[html:y945]{lexer_I_945}
                        |  \hyperref[html:y945]{lexer_I_945}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y810}
NA_KU_810               :  \hyperref[html:y950]{lexer_J_950}
                        |  \hyperref[html:y950]{lexer_J_950}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y811}
I_BO_811                :  \hyperref[html:y955]{lexer_K_955}
                        |  \hyperref[html:y955]{lexer_K_955}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y812}
number_812              :  \hyperref[html:y960]{lexer_L_960}
                        ;

\phantomsection\label{html:y813}
GIhEK_BO_813            :  \hyperref[html:y965]{lexer_M_965}
                        |  \hyperref[html:y965]{lexer_M_965}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y814}
GIhEK_KE_814            :  \hyperref[html:y966]{lexer_N_966}
                        |  \hyperref[html:y966]{lexer_N_966}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y815}
tense_modal_815         :  \hyperref[html:y970]{lexer_O_970}
                        |  \hyperref[html:y970]{lexer_O_970}  \hyperref[html:y32]{free_modifier_32}
                        |  \hyperref[html:y437]{FIhO_437}  \hyperref[html:y130]{selbri_130}  \hyperref[html:y458]{FEhU_gap_458}
                        ;

\phantomsection\label{html:y816}
GIK_816                 :  \hyperref[html:y980]{lexer_P_980}
                        |  \hyperref[html:y980]{lexer_P_980}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y817}
lerfu_string_817        :  \hyperref[html:y985]{lexer_Q_985}
                        ;

\phantomsection\label{html:y818}
GIhEK_818               :  \hyperref[html:y990]{lexer_R_990}
                        |  \hyperref[html:y990]{lexer_R_990}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y819}
I_819                   :  \hyperref[html:y995]{lexer_S_995}
                        |  \hyperref[html:y995]{lexer_S_995}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y820}
I_JEK_820               :  \hyperref[html:y1000]{lexer_T_1000}
                        |  \hyperref[html:y1000]{lexer_T_1000}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y821}
JEK_BO_821              :  \hyperref[html:y1005]{lexer_U_1005}
                        |  \hyperref[html:y1005]{lexer_U_1005}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y822}
JOIK_BO_822             :  \hyperref[html:y1010]{lexer_V_1010}
                        |  \hyperref[html:y1010]{lexer_V_1010}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y823}
JOIK_KE_823             :  \hyperref[html:y1015]{lexer_W_1015}
                        |  \hyperref[html:y1015]{lexer_W_1015}  \hyperref[html:y32]{free_modifier_32}
                        ;

\phantomsection\label{html:y824}
PA_MOI_824              :  \hyperref[html:y1025]{lexer_Y_1025}
                        ;

/* The following rules are used only in lexer processing.  They have been
   tested for ambiguity at various levels in the YACC grammar, but are in
   the recursive descent lexer in the current parser.  The lexer inserts
   the lexer tokens before the processed strings, but leaves the original
   tokens.  */

\phantomsection\label{html:y905}
lexer_A_905             :  \hyperref[html:y701]{lexer_A_701}  \hyperref[html:y906]{utt_ordinal_root_906}
                        ;

\phantomsection\label{html:y906}
utt_ordinal_root_906    :  \hyperref[html:y986]{lerfu_string_root_986}  \hyperref[html:y661]{MAI_661}
                        |  \hyperref[html:y961]{number_root_961}  \hyperref[html:y661]{MAI_661}
                        ;

\phantomsection\label{html:y910}
lexer_B_910             :  \hyperref[html:y702]{lexer_B_702}  \hyperref[html:y911]{EK_root_911}
                        ;

\phantomsection\label{html:y911}
EK_root_911             :  \hyperref[html:y501]{A_501}
                        |  \hyperref[html:y596]{SE_596}  \hyperref[html:y501]{A_501}
                        |  \hyperref[html:y578]{NA_578}  \hyperref[html:y501]{A_501}
                        |  \hyperref[html:y501]{A_501}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y596]{SE_596}  \hyperref[html:y501]{A_501}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y578]{NA_578}  \hyperref[html:y501]{A_501}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y578]{NA_578}  \hyperref[html:y596]{SE_596}  \hyperref[html:y501]{A_501}
                        |  \hyperref[html:y578]{NA_578}  \hyperref[html:y596]{SE_596}  \hyperref[html:y501]{A_501}  \hyperref[html:y581]{NAI_581}
                        ;

\phantomsection\label{html:y915}
lexer_C_915             :  \hyperref[html:y703]{lexer_C_703}  \hyperref[html:y911]{EK_root_911}  \hyperref[html:y508]{BO_508}
                        |  \hyperref[html:y703]{lexer_C_703}  \hyperref[html:y911]{EK_root_911}  \hyperref[html:y971]{simple_tag_971}  \hyperref[html:y508]{BO_508}
                        ;

\phantomsection\label{html:y916}
lexer_D_916             :  \hyperref[html:y704]{lexer_D_704}  \hyperref[html:y911]{EK_root_911}  \hyperref[html:y551]{KE_551}
                        |  \hyperref[html:y704]{lexer_D_704}  \hyperref[html:y911]{EK_root_911}  \hyperref[html:y971]{simple_tag_971}  \hyperref[html:y551]{KE_551}
                        ;

\phantomsection\label{html:y925}
lexer_E_925             :  \hyperref[html:y705]{lexer_E_705}  \hyperref[html:y926]{JEK_root_926}
                        ;

\phantomsection\label{html:y926}
JEK_root_926            :  \hyperref[html:y546]{JA_546}
                        |  \hyperref[html:y546]{JA_546}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y578]{NA_578}  \hyperref[html:y546]{JA_546}
                        |  \hyperref[html:y578]{NA_578}  \hyperref[html:y546]{JA_546}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y596]{SE_596}  \hyperref[html:y546]{JA_546}
                        |  \hyperref[html:y596]{SE_596}  \hyperref[html:y546]{JA_546}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y578]{NA_578}  \hyperref[html:y596]{SE_596}  \hyperref[html:y546]{JA_546}
                        |  \hyperref[html:y578]{NA_578}  \hyperref[html:y596]{SE_596}  \hyperref[html:y546]{JA_546}  \hyperref[html:y581]{NAI_581}
                        ;

\phantomsection\label{html:y930}
lexer_F_930             :  \hyperref[html:y706]{lexer_F_706}  \hyperref[html:y931]{JOIK_root_931}
                        ;

\phantomsection\label{html:y931}
JOIK_root_931           :  \hyperref[html:y548]{JOI_548}
                        |  \hyperref[html:y548]{JOI_548}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y596]{SE_596}  \hyperref[html:y548]{JOI_548}
                        |  \hyperref[html:y596]{SE_596}  \hyperref[html:y548]{JOI_548}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y932]{interval_932}
                        |  \hyperref[html:y656]{GAhO_656}  \hyperref[html:y932]{interval_932}  \hyperref[html:y656]{GAhO_656}
                        ;

\phantomsection\label{html:y932}
interval_932            :  \hyperref[html:y507]{BIhI_507}
                        |  \hyperref[html:y507]{BIhI_507}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y596]{SE_596}  \hyperref[html:y507]{BIhI_507}
                        |  \hyperref[html:y596]{SE_596}  \hyperref[html:y507]{BIhI_507}  \hyperref[html:y581]{NAI_581}
                        ;

\phantomsection\label{html:y935}
lexer_G_935             :  \hyperref[html:y707]{lexer_G_707}  \hyperref[html:y537]{GA_537}
                        |  \hyperref[html:y707]{lexer_G_707}  \hyperref[html:y596]{SE_596}  \hyperref[html:y537]{GA_537}
                        |  \hyperref[html:y707]{lexer_G_707}  \hyperref[html:y537]{GA_537}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y707]{lexer_G_707}  \hyperref[html:y596]{SE_596}  \hyperref[html:y537]{GA_537}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y707]{lexer_G_707}  \hyperref[html:y971]{simple_tag_971}  \hyperref[html:y981]{GIK_root_981}
                        |  \hyperref[html:y707]{lexer_G_707}  \hyperref[html:y931]{JOIK_root_931}  \hyperref[html:y539]{GI_539}
                        ;

\phantomsection\label{html:y940}
lexer_H_940             :  \hyperref[html:y708]{lexer_H_708}  \hyperref[html:y544]{GUhA_544}
                        |  \hyperref[html:y708]{lexer_H_708}  \hyperref[html:y596]{SE_596}  \hyperref[html:y544]{GUhA_544}
                        |  \hyperref[html:y708]{lexer_H_708}  \hyperref[html:y544]{GUhA_544}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y708]{lexer_H_708}  \hyperref[html:y596]{SE_596}  \hyperref[html:y544]{GUhA_544}  \hyperref[html:y581]{NAI_581}
                        ;

\phantomsection\label{html:y945}
lexer_I_945             :  \hyperref[html:y709]{lexer_I_709}  \hyperref[html:y583]{NAhE_583}  \hyperref[html:y508]{BO_508}
                        ;

\phantomsection\label{html:y950}
lexer_J_950             :  \hyperref[html:y710]{lexer_J_710}  \hyperref[html:y578]{NA_578}  \hyperref[html:y556]{KU_556}
                        ;

\phantomsection\label{html:y955}
lexer_K_955             :  \hyperref[html:y711]{lexer_K_711}  \hyperref[html:y956]{I_root_956}  \hyperref[html:y508]{BO_508}
                        |  \hyperref[html:y711]{lexer_K_711}  \hyperref[html:y956]{I_root_956}  \hyperref[html:y971]{simple_tag_971}  \hyperref[html:y508]{BO_508}
                        ;

\phantomsection\label{html:y956}
I_root_956              :  \hyperref[html:y545]{I_545}
                        |  \hyperref[html:y545]{I_545}  \hyperref[html:y957]{simple_JOIK_JEK_957}
                        ;

\phantomsection\label{html:y957}
simple_JOIK_JEK_957     :  \hyperref[html:y806]{JOIK_806}
                        |  \hyperref[html:y805]{JEK_805}
                        ;
                        /* no freemod in this version; cf. \hyperref[html:y422]{JOIK_JEK_422} */
                        /* this reference to a version of JOIK and JEK
                           which already have the lexer tokens attached
                           prevents shift/reduce errors.  The problem is
                           resolved in a hard-coded parser implementation
                           which builds lexer_K, before lexer_S, before
                           lexer_E and lexer_F. */

\phantomsection\label{html:y960}
lexer_L_960             :  \hyperref[html:y712]{lexer_L_712}  \hyperref[html:y961]{number_root_961}
                        ;

\phantomsection\label{html:y961}
number_root_961         :  \hyperref[html:y672]{PA_672}
                        |  \hyperref[html:y961]{number_root_961}  \hyperref[html:y672]{PA_672}
                        |  \hyperref[html:y961]{number_root_961}  \hyperref[html:y987]{lerfu_word_987}
                        ;

\phantomsection\label{html:y965}
lexer_M_965             :  \hyperref[html:y713]{lexer_M_713}  \hyperref[html:y991]{GIhEK_root_991}  \hyperref[html:y508]{BO_508}
                        |  \hyperref[html:y713]{lexer_M_713}  \hyperref[html:y991]{GIhEK_root_991}  \hyperref[html:y971]{simple_tag_971}  \hyperref[html:y508]{BO_508}
                        ;

\phantomsection\label{html:y966}
lexer_N_966             :  \hyperref[html:y714]{lexer_N_714}  \hyperref[html:y991]{GIhEK_root_991}  \hyperref[html:y551]{KE_551}
                        |  \hyperref[html:y714]{lexer_N_714}  \hyperref[html:y991]{GIhEK_root_991}  \hyperref[html:y971]{simple_tag_971}  \hyperref[html:y551]{KE_551}
                        ;

\phantomsection\label{html:y970}
lexer_O_970             :  \hyperref[html:y715]{lexer_O_715}  \hyperref[html:y972]{simple_tense_modal_972}
                        ;
/* the following rule is a lexer version of non-terminal _815 for compounding
   PU/modals; it disallows the lexer picking out FIhO clauses, which would
   require it to have knowledge of the main parser grammar */

\phantomsection\label{html:y971}
simple_tag_971          :  \hyperref[html:y972]{simple_tense_modal_972}
                        |  \hyperref[html:y971]{simple_tag_971}  \hyperref[html:y957]{simple_JOIK_JEK_957}
                                \hyperref[html:y972]{simple_tense_modal_972}
                        ;

\phantomsection\label{html:y972}
simple_tense_modal_972  :  \hyperref[html:y973]{simple_tense_modal_A_973}
                        |  \hyperref[html:y583]{NAhE_583}  \hyperref[html:y973]{simple_tense_modal_A_973}
                        |  \hyperref[html:y554]{KI_554}
                        |  \hyperref[html:y522]{CUhE_522}
                        ;

\phantomsection\label{html:y973}
simple_tense_modal_A_973:  \hyperref[html:y974]{modal_974}
                        |  \hyperref[html:y974]{modal_974}  \hyperref[html:y554]{KI_554}
                        |  \hyperref[html:y977]{tense_A_977}
                        ;

\phantomsection\label{html:y974}
modal_974               :  \hyperref[html:y975]{modal_A_975}
                        |  \hyperref[html:y975]{modal_A_975}  \hyperref[html:y581]{NAI_581}
                        ;

\phantomsection\label{html:y975}
modal_A_975             :  \hyperref[html:y502]{BAI_502}
                        |  \hyperref[html:y596]{SE_596}  \hyperref[html:y502]{BAI_502}
                        ;

\phantomsection\label{html:y977}
tense_A_977             :  \hyperref[html:y978]{tense_B_978}
                        |  \hyperref[html:y978]{tense_B_978}  \hyperref[html:y554]{KI_554}
                        ;

\phantomsection\label{html:y978}
tense_B_978             :  \hyperref[html:y979]{tense_C_979}
                        |  \hyperref[html:y514]{CAhA_514}
                        |  \hyperref[html:y979]{tense_C_979}  \hyperref[html:y514]{CAhA_514}
                        ;
/* specifies actuality/potentiality of the bridi */

/* puca'a = actually was */
/* baca'a = actually will be */
/* bapu'i = can and will have */
/* banu'o = can, but won't have yet */
/* canu'ojebapu'i = can, hasn't yet, but will */

\phantomsection\label{html:y979}
tense_C_979             :  \hyperref[html:y1030]{time_1030}
   /* time-only */
   /* space defaults to time-space reference space */

                    |  \hyperref[html:y1040]{space_1040}
   /* can include time if specified with VIhA */
   /* otherwise time defaults to the time-space reference time */

                    |  \hyperref[html:y1030]{time_1030}  \hyperref[html:y1040]{space_1040}
   /* time and space - If \hyperref[html:y1040]{space_1040} is marked with
   VIhA for space-time the tense may be self-contradictory */
   /* interval prop before space_time is for time distribution */
                        |  \hyperref[html:y1040]{space_1040}  \hyperref[html:y1030]{time_1030}
                        ;

\phantomsection\label{html:y980}
lexer_P_980             :  \hyperref[html:y716]{lexer_P_716}  \hyperref[html:y981]{GIK_root_981}
                        ;

\phantomsection\label{html:y981}
GIK_root_981            :  \hyperref[html:y539]{GI_539}
                        |  \hyperref[html:y539]{GI_539}  \hyperref[html:y581]{NAI_581}
                        ;

\phantomsection\label{html:y985}
lexer_Q_985             :  \hyperref[html:y717]{lexer_Q_717}  \hyperref[html:y986]{lerfu_string_root_986}
                        ;

\phantomsection\label{html:y986}
lerfu_string_root_986   :  \hyperref[html:y987]{lerfu_word_987}
                        |  \hyperref[html:y986]{lerfu_string_root_986}  \hyperref[html:y987]{lerfu_word_987}
                        |  \hyperref[html:y986]{lerfu_string_root_986}  \hyperref[html:y672]{PA_672}
                        ;

\phantomsection\label{html:y987}
lerfu_word_987          :  \hyperref[html:y513]{BY_513}
                        |  \hyperref[html:y559]{LAU_559}  \hyperref[html:y987]{lerfu_word_987}
                        |  \hyperref[html:y605]{TEI_605}  \hyperref[html:y986]{lerfu_string_root_986}  \hyperref[html:y533]{FOI_533}
                        ;

\phantomsection\label{html:y990}
lexer_R_990             :  \hyperref[html:y718]{lexer_R_718}  \hyperref[html:y991]{GIhEK_root_991}
                        ;

\phantomsection\label{html:y991}
GIhEK_root_991          :  \hyperref[html:y541]{GIhA_541}
                        |  \hyperref[html:y596]{SE_596}  \hyperref[html:y541]{GIhA_541}
                        |  \hyperref[html:y578]{NA_578}  \hyperref[html:y541]{GIhA_541}
                        |  \hyperref[html:y541]{GIhA_541}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y596]{SE_596}  \hyperref[html:y541]{GIhA_541}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y578]{NA_578}  \hyperref[html:y541]{GIhA_541}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y578]{NA_578}  \hyperref[html:y596]{SE_596}  \hyperref[html:y541]{GIhA_541}
                        |  \hyperref[html:y578]{NA_578}  \hyperref[html:y596]{SE_596}  \hyperref[html:y541]{GIhA_541}  \hyperref[html:y581]{NAI_581}
                        ;

\phantomsection\label{html:y995}
lexer_S_995             :  \hyperref[html:y719]{lexer_S_719}  \hyperref[html:y545]{I_545}
                        ;

\phantomsection\label{html:y1000}
lexer_T_1000            :  \hyperref[html:y720]{lexer_T_720}  \hyperref[html:y545]{I_545}  \hyperref[html:y957]{simple_JOIK_JEK_957}
                        ;

\phantomsection\label{html:y1005}
lexer_U_1005            :  \hyperref[html:y721]{lexer_U_721}  \hyperref[html:y926]{JEK_root_926}  \hyperref[html:y508]{BO_508}
                        |  \hyperref[html:y721]{lexer_U_721}  \hyperref[html:y926]{JEK_root_926}  \hyperref[html:y971]{simple_tag_971}  \hyperref[html:y508]{BO_508}
                        ;

\phantomsection\label{html:y1010}
lexer_V_1010            :  \hyperref[html:y722]{lexer_V_722}  \hyperref[html:y931]{JOIK_root_931}  \hyperref[html:y508]{BO_508}
                        |  \hyperref[html:y722]{lexer_V_722}  \hyperref[html:y931]{JOIK_root_931}  \hyperref[html:y971]{simple_tag_971}  \hyperref[html:y508]{BO_508}
                        ;

\phantomsection\label{html:y1015}
lexer_W_1015            :  \hyperref[html:y723]{lexer_W_723}  \hyperref[html:y931]{JOIK_root_931}  \hyperref[html:y551]{KE_551}
                        |  \hyperref[html:y723]{lexer_W_723}  \hyperref[html:y931]{JOIK_root_931}  \hyperref[html:y971]{simple_tag_971}  \hyperref[html:y551]{KE_551}
                        ;

\phantomsection\label{html:y1025}
lexer_Y_1025            :  \hyperref[html:y725]{lexer_Y_725}  \hyperref[html:y961]{number_root_961}  \hyperref[html:y663]{MOI_663}
                        |  \hyperref[html:y725]{lexer_Y_725}  \hyperref[html:y986]{lerfu_string_root_986}  \hyperref[html:y663]{MOI_663}
                        ;

\phantomsection\label{html:y1030}
time_1030               :  \hyperref[html:y624]{ZI_624}
                        |  \hyperref[html:y624]{ZI_624}  \hyperref[html:y1031]{time_A_1031}
                        |  \hyperref[html:y1031]{time_A_1031}
                        ;

\phantomsection\label{html:y1031}
time_A_1031             :  \hyperref[html:y1032]{time_B_1032}
                        |  \hyperref[html:y1034]{time_interval_1034}
                        |  \hyperref[html:y1032]{time_B_1032}  \hyperref[html:y1034]{time_interval_1034}
                        ;

\phantomsection\label{html:y1032}
time_B_1032             :  \hyperref[html:y1033]{time_offset_1033}
                        |  \hyperref[html:y1032]{time_B_1032}  \hyperref[html:y1033]{time_offset_1033}
                        ;

\phantomsection\label{html:y1033}
time_offset_1033        :  \hyperref[html:y1035]{time_direction_1035}
                        |  \hyperref[html:y1035]{time_direction_1035}  \hyperref[html:y624]{ZI_624}
                        ;

\phantomsection\label{html:y1034}
time_interval_1034      :  \hyperref[html:y622]{ZEhA_622}
                        |  \hyperref[html:y622]{ZEhA_622}  \hyperref[html:y1035]{time_direction_1035}
                        |  \hyperref[html:y1036]{time_int_props_1036}
                        |  \hyperref[html:y622]{ZEhA_622}  \hyperref[html:y1036]{time_int_props_1036}
                        |  \hyperref[html:y622]{ZEhA_622}  \hyperref[html:y1035]{time_direction_1035}  \hyperref[html:y1036]{time_int_props_1036}
                        ;

\phantomsection\label{html:y1035}
time_direction_1035     :  \hyperref[html:y592]{PU_592}
                        |  \hyperref[html:y592]{PU_592}  \hyperref[html:y581]{NAI_581}
                        ;

\phantomsection\label{html:y1036}
time_int_props_1036     :  \hyperref[html:y1051]{interval_property_1051}
                        |  \hyperref[html:y1036]{time_int_props_1036}  \hyperref[html:y1051]{interval_property_1051}
                        ;

\phantomsection\label{html:y1040}
space_1040              :  \hyperref[html:y1042]{space_A_1042}
                        |  \hyperref[html:y1041]{space_motion_1041}
                        |  \hyperref[html:y1042]{space_A_1042}  \hyperref[html:y1041]{space_motion_1041}
                        ;

\phantomsection\label{html:y1041}
space_motion_1041       :  \hyperref[html:y577]{MOhI_577}  \hyperref[html:y1045]{space_offset_1045}
                        ;

\phantomsection\label{html:y1042}
space_A_1042            :  \hyperref[html:y613]{VA_613}
                        |  \hyperref[html:y613]{VA_613}  \hyperref[html:y1043]{space_B_1043}
                        |  \hyperref[html:y1043]{space_B_1043}
                        ;

\phantomsection\label{html:y1043}
space_B_1043            :  \hyperref[html:y1044]{space_C_1044}
                        |  \hyperref[html:y1046]{space_intval_1046}
                        |  \hyperref[html:y1044]{space_C_1044}  \hyperref[html:y1046]{space_intval_1046}
                        ;

\phantomsection\label{html:y1044}
space_C_1044            :  \hyperref[html:y1045]{space_offset_1045}
                        |  \hyperref[html:y1044]{space_C_1044}  \hyperref[html:y1045]{space_offset_1045}
                        ;

\phantomsection\label{html:y1045}
space_offset_1045       :  \hyperref[html:y1048]{space_direction_1048}
                        |  \hyperref[html:y1048]{space_direction_1048}  \hyperref[html:y613]{VA_613}
                        ;

\phantomsection\label{html:y1046}
space_intval_1046       :  \hyperref[html:y1047]{space_intval_A_1047}
                        |  \hyperref[html:y1047]{space_intval_A_1047}  \hyperref[html:y1048]{space_direction_1048}
                        |  \hyperref[html:y1049]{space_int_props_1049}
                        |  \hyperref[html:y1047]{space_intval_A_1047}  \hyperref[html:y1049]{space_int_props_1049}
                        |  \hyperref[html:y1047]{space_intval_A_1047}  \hyperref[html:y1048]{space_direction_1048}
                                  \hyperref[html:y1049]{space_int_props_1049}
                        ;

\phantomsection\label{html:y1047}
space_intval_A_1047     :  \hyperref[html:y615]{VEhA_615}
                        |  \hyperref[html:y616]{VIhA_616}
                        |  \hyperref[html:y615]{VEhA_615}  \hyperref[html:y616]{VIhA_616}
                        ;

\phantomsection\label{html:y1048}
space_direction_1048    :  \hyperref[html:y528]{FAhA_528}
                        |  \hyperref[html:y528]{FAhA_528}  \hyperref[html:y581]{NAI_581}
                        ;

\phantomsection\label{html:y1049}
space_int_props_1049    :  \hyperref[html:y1050]{space_int_props_A_1050}
                        |  \hyperref[html:y1049]{space_int_props_1049}  \hyperref[html:y1050]{space_int_props_A_1050}
                        ;

\phantomsection\label{html:y1050}
space_int_props_A_1050  : \hyperref[html:y530]{FEhE_530}  \hyperref[html:y1051]{interval_property_1051}
                        ;

/* This terminal gives an interval size in space-time (VEhA), and possibly a
   dimensionality of the interval.  The dimensionality may also be used
   with the interval size left unspecified.  When this terminal is used for the
   spacetime origin, then barring any overriding VIhA, a VIhA here defines
   the dimensionality of the space-time being discussed.                 */

\phantomsection\label{html:y1051}
interval_property_1051  :  \hyperref[html:y961]{number_root_961}  \hyperref[html:y594]{ROI_594}
                        |  \hyperref[html:y961]{number_root_961}  \hyperref[html:y594]{ROI_594}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y604]{TAhE_604}
                        |  \hyperref[html:y604]{TAhE_604}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y621]{ZAhO_621}
                        |  \hyperref[html:y621]{ZAhO_621}  \hyperref[html:y581]{NAI_581}
                        ;

/* extensional/intensional interval parameters */
/* These may be appended to any defined interval, or may stand in place of
   either time or space tenses.  If no other tense is present, this terminal
   stands for the time-space interval parameter with an unspecified interval.*/

   /* roroi = always and everywhere */
   /* roroiku'avi = always here (ku'a = intersection) */
   /* puroroi = always in the past
   /* paroi = once upon a time (somewhere) */
   /* paroiku'avi = once upon a time here */

/* The following are \q{Lexer-only rules}, covered by steps 1-4 described
   at the beginning.  The grammar of these constructs is nonexistent,
   except possibly in cases where they interact with each other.  Even
   there, however, the effects are semantic rather than grammatical.  Where
   it is believed possible that conflicts could exist, the grammar of these
   constructs has been put in the above grammar, even though the
   lexer/Preparser will actually prevent these from being passed thru to
   the parse routine.  (Otherwise we have to put unacceptably fancy code in
   the PreParser to determine just when these can be passed thru, and when
   they can't.)  Constructs in this category include quotes and indicators
   as defined above.  (The above grammar handles utterance scope
   (free_modifier) and clause scope (gap) applications of the latter,
   however, and indicators should be allowed to be absorbed into almost any
   word without changing its grammar.

   \hyperref[html:y601]{SI_601}, \hyperref[html:y595]{SA_595}, and \hyperref[html:y603]{SU_603} are metalinguistic erasers.

\phantomsection\label{html:y1100}
token_1100              :  \hyperref[html:y698]{any_word_698}
                        |  \hyperref[html:y503]{BAhE_503}  \hyperref[html:y698]{any_word_698}
                        |  \hyperref[html:y699]{anything_699}
                        |  \hyperref[html:y698]{any_word_698}  \hyperref[html:y511]{BU_511}
                        |  \hyperref[html:y698]{any_word_698}  \hyperref[html:y524]{DAhO_524}
                        |  \hyperref[html:y698]{any_word_698}  \hyperref[html:y536]{FUhO_536}
                        |  \hyperref[html:y698]{any_word_698}  \hyperref[html:y535]{FUhE_535}
                        |  \hyperref[html:y698]{any_word_698}  \hyperref[html:y612]{UI_612}
                        |  \hyperref[html:y698]{any_word_698}  \hyperref[html:y612]{UI_612}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y698]{any_word_698}  \hyperref[html:y619]{Y_619}
                        |  \hyperref[html:y698]{any_word_698}  \hyperref[html:y515]{CAI_515}
                        |  \hyperref[html:y698]{any_word_698}  \hyperref[html:y515]{CAI_515}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y612]{UI_612}  \hyperref[html:y581]{NAI_581}
                        |  \hyperref[html:y515]{CAI_515}  \hyperref[html:y581]{NAI_581}
                        ;

\phantomsection\label{html:y1101}
null_1101               :  \hyperref[html:y698]{any_word_698}  \hyperref[html:y601]{SI_601}
                        |  possibly_unlexable_word  (PAUSE)  \hyperref[html:y601]{SI_601}
                        |  \hyperref[html:y20]{utterance_20}  \hyperref[html:y595]{SA_595}
                        |  possibly unlexable string  (PAUSE)  \hyperref[html:y595]{SA_595}
                           erases back to the last individual token
                           I or NIhO or start of text, ignoring the
                           insides of ZOI, ZO, and LOhU/LEhU quotes.
                           Start of text is defined for SU below.
                        |  \hyperref[html:y3]{text_C_3}  \hyperref[html:y603]{SU_603}
                        |  possibly unparsable text  (PAUSE)  \hyperref[html:y603]{SU_603}
                           erases back to start of text which is the
                           beginning of a speaker's statement,
                           a parenthesis (TO/TOI), a LU/LIhU quote,
                           or a TUhE/TUhU utterance string.
                        ;

*/
%%



\sect{EBNF Grammar of Lojban}
Lojban Machine Grammar, EBNF Version, 3rd Baseline as of 10
    January 1997 

This document is explicitly dedicated to the public domain
    by its author, the Logical Language Group Inc. Contact that
    organization at: 2904 Beau Lane, Fairfax VA 22031 USA
    703-385-0273 (intl: +1 703 385 0273)

Explanation of notation: All rules have the form:

<div class="c2">
      name<sub>number</sub> = bnf-expression
    </div>
    <br />
    <br />
     which means that the grammatical construct \q{name} is defined
    by \q{bnf-expression}. The number cross-references this grammar
    with the rule numbers in the YACC grammar. The names are the
    same as those in the YACC grammar, except that subrules are
    labeled with A, B, C, ... in the YACC grammar and with 1, 2, 3,
    ... in this grammar. In addition, rule 971 is \q{simple\_tag} in
    the YACC grammar but \q{stag} in this grammar, because of its
    frequent appearance. 
\begin{enumerate}
\item Names in lower case are grammatical constructs.
\item Names in UPPER CASE are selma'o (lexeme) names, and are terminals.
\item Concatenation is expressed by juxtaposition with no operator symbol.
\item | represents alternation (choice).
\item [] represents an optional element.
\item \&represents and/or (\q{A \&B} is the same as ``A | B | A B'').
\item ... represents optional repetition of the construct to the left. Left-grouping is implied; right-grouping is shown by explicit self-referential recursion with no \q{...}
\item () serves to indicate the grouping of the other operators. Otherwise, \q{...} binds closer than \& which binds closer than |.
\item # is shorthand for \q{[free ...]}, a construct which appears in many places.
\item // encloses an elidable terminator, which may be omitted (without change of meaning) if no grammatical ambiguity results. \end{enumerate}\begin{description} <dd>\phantomsection\label{html:b0}
\hyperref[html:y0]{}</dd>
\item[text<sub>0</sub> =] [NAI ...] [CMENE ... # | (indicators \&free ...)] [joik-jek] text-1 \phantomsection\label{html:b2}
\hyperref[html:y2]{}
\item[text-1<sub>2</sub> =] [(I [jek | joik] [[stag] BO] #) ... | NIhO ... #] [paragraphs] \phantomsection\label{html:b4}
\hyperref[html:y4]{}
\item[paragraphs<sub>4</sub> =] paragraph [NIhO ... # paragraphs] \phantomsection\label{html:b10}
\hyperref[html:y10]{}
\item[paragraph<sub>10</sub> =] (statement | fragment) [I # [statement | fragment]] ... \phantomsection\label{html:b11}
\hyperref[html:y11]{}
\item[statement<sub>11</sub> =] statement-1<br /> | prenex statement \phantomsection\label{html:b12}
\hyperref[html:y12]{}
\item[statement-1<sub>12</sub> =] statement-2 [I joik-jek [statement-2]] ... \phantomsection\label{html:b13}
\hyperref[html:y13]{}
\item[statement-2<sub>13</sub> =] statement-3 [I [jek | joik] [stag] BO # [statement-2]] \phantomsection\label{html:b14}
\hyperref[html:y14]{}
\item[statement-3<sub>14</sub> =] sentence<br /> | [tag] TUhE # text-1 /TUhU#/ \phantomsection\label{html:b20}
\hyperref[html:y20]{}
\item[fragment<sub>20</sub> =] ek #<br /> | gihek #<br /> | quantifier<br /> | NA #<br /> | terms /VAU#/<br /> | prenex<br /> | relative-clauses<br /> | links<br /> | linkargs \phantomsection\label{html:b30}
\hyperref[html:y30]{}
\item[prenex<sub>30</sub> =] terms ZOhU # \phantomsection\label{html:b40}
\hyperref[html:y40]{}
\item[sentence<sub>40</sub> =] [terms [CU #]] bridi-tail \phantomsection\label{html:b41}
\hyperref[html:y41]{}
\item[subsentence<sub>41</sub> =] sentence<br /> | prenex subsentence \phantomsection\label{html:b50}
\hyperref[html:y50]{}
\item[bridi-tail<sub>50</sub> =] bridi-tail-1 [gihek [stag] KE # bridi-tail /KEhE#/ tail-terms] \phantomsection\label{html:b51}
\hyperref[html:y51]{}
\item[bridi-tail-1<sub>51</sub> =] bridi-tail-2 [gihek # bridi-tail-2 tail-terms] ... \phantomsection\label{html:b52}
\hyperref[html:y52]{}
\item[bridi-tail-2<sub>52</sub> =] bridi-tail-3 [gihek [stag] BO # bridi-tail-2 tail-terms] \phantomsection\label{html:b53}
\hyperref[html:y53]{}
\item[bridi-tail-3<sub>53</sub> =] selbri tail-terms<br /> | gek-sentence \phantomsection\label{html:b54}
\hyperref[html:y54]{}
\item[gek-sentence<sub>54</sub> =] gek subsentence gik subsentence tail-terms<br /> | [tag] KE # gek-sentence /KEhE#/<br /> | NA # gek-sentence \phantomsection\label{html:b71}
\hyperref[html:y71]{}
\item[tail-terms<sub>71</sub> =] [terms] /VAU#/ \phantomsection\label{html:b80}
\hyperref[html:y80]{}
\item[terms<sub>80</sub> =] terms-1 ... \phantomsection\label{html:b81}
\hyperref[html:y81]{}
\item[terms-1<sub>81</sub> =] terms-2 [PEhE # joik-jek terms-2] ... \phantomsection\label{html:b82}
\hyperref[html:y82]{}
\item[terms-2<sub>82</sub> =] term [CEhE # term] ... \phantomsection\label{html:b83}
\hyperref[html:y83]{}
\item[term<sub>83</sub> =] sumti | (tag | FA #) (sumti<br /> | /KU#/)<br /> | termset<br /> | NA KU # \phantomsection\label{html:b85}
\hyperref[html:y85]{}
\item[termset<sub>85</sub> =] NUhI # gek terms /NUhU#/ gik terms /NUhU#/<br /> | NUhI # terms /NUhU#/ \phantomsection\label{html:b90}
\hyperref[html:y90]{}
\item[sumti<sub>90</sub> =] sumti-1 [VUhO # relative-clauses] \phantomsection\label{html:b91}
\hyperref[html:y91]{}
\item[sumti-1<sub>91</sub> =] sumti-2 [(ek | joik) [stag] KE # sumti /KEhE#/] \phantomsection\label{html:b92}
\hyperref[html:y92]{}
\item[sumti-2<sub>92</sub> =] sumti-3 [joik-ek sumti-3] ... \phantomsection\label{html:b93}
\hyperref[html:y93]{}
\item[sumti-3<sub>93</sub> =] sumti-4 [(ek<br /> | joik) [stag] BO # sumti-3] \phantomsection\label{html:b94}
\hyperref[html:y94]{}
\item[sumti-4<sub>94</sub> =] sumti-5<br /> | gek sumti gik sumti-4 \phantomsection\label{html:b95}
\hyperref[html:y95]{}
\item[sumti-5<sub>95</sub> =] [quantifier] sumti-6 [relative-clauses]<br /> | quantifier selbri /KU#/ [relative-clauses] \phantomsection\label{html:b97}
\hyperref[html:y97]{}
\item[sumti-6<sub>97</sub> =] (LAhE # | NAhE BO #) [relative-clauses] sumti /LUhU#/<br /> | KOhA #<br /> | lerfu-string /BOI#/<br /> | LA # [relative-clauses] CMENE ... #<br /> | (LA | LE) # sumti-tail /KU#/<br /> | LI # mex /LOhO#/<br /> | ZO any-word #<br /> | LU text /LIhU#/<br /> | LOhU any-word ... LEhU #<br /> | ZOI any-word anything any-word # \phantomsection\label{html:b111}
\hyperref[html:y111]{}
\item[sumti-tail<sub>111</sub> =] [sumti-6 [relative-clauses]] sumti-tail-1<br /> | relative-clauses sumti-tail-1 \phantomsection\label{html:b112}
\hyperref[html:y112]{}
\item[sumti-tail-1<sub>112</sub> =] [quantifier] selbri [relative-clauses]<br /> | quantifier sumti \phantomsection\label{html:b121}
\hyperref[html:y121]{}
\item[relative-clauses<sub>121</sub> =] relative-clause [ZIhE # relative-clause] ... \phantomsection\label{html:b122}
\hyperref[html:y122]{}
\item[relative-clause<sub>122</sub> =] GOI # term /GEhU#/<br /> | NOI # subsentence /KUhO#/ \phantomsection\label{html:b130}
\hyperref[html:y130]{}
\item[selbri<sub>130</sub> =] [tag] selbri-1 \phantomsection\label{html:b131}
\hyperref[html:y131]{}
\item[selbri-1<sub>131</sub> =] selbri-2<br /> | NA # selbri \phantomsection\label{html:b132}
\hyperref[html:y132]{}
\item[selbri-2<sub>132</sub> =] selbri-3 [CO # selbri-2] \phantomsection\label{html:b133}
\hyperref[html:y133]{}
\item[selbri-3<sub>133</sub> =] selbri-4 ... \phantomsection\label{html:b134}
\hyperref[html:y134]{}
\item[selbri-4<sub>134</sub> =] selbri-5 [joik-jek selbri-5 | joik [stag] KE # selbri-3 /KEhE#/] ... \phantomsection\label{html:b135}
\hyperref[html:y135]{}
\item[selbri-5<sub>135</sub> =] selbri-6 [(jek | joik) [stag] BO # selbri-5] \phantomsection\label{html:b136}
\hyperref[html:y136]{}
\item[selbri-6<sub>136</sub> =] tanru-unit [BO # selbri-6]<br /> | [NAhE #] guhek selbri gik selbri-6 \phantomsection\label{html:b150}
\hyperref[html:y150]{}
\item[tanru-unit<sub>150</sub> =] tanru-unit-1 [CEI # tanru-unit-1] ... \phantomsection\label{html:b151}
\hyperref[html:y151]{}
\item[tanru-unit-1<sub>151</sub> =] tanru-unit-2 [linkargs] \phantomsection\label{html:b152}
\hyperref[html:y152]{}
\item[tanru-unit-2<sub>152</sub> =] BRIVLA #<br /> | GOhA [RAhO] #<br /> | KE # selbri-3 /KEhE#/<br /> | ME # sumti /MEhU#/ [MOI #] | (number<br /> | lerfu-string) MOI #<br /> | NUhA # mex-operator<br /> | SE # tanru-unit-2<br /> | JAI # [tag] tanru-unit-2<br /> | any-word (ZEI any-word) ...<br /> | NAhE # tanru-unit-2<br /> | NU [NAI] # [joik-jek NU [NAI] #] ... subsentence /KEI#/ \phantomsection\label{html:b160}
\hyperref[html:y160]{}
\item[linkargs<sub>160</sub> =] BE # term [links] /BEhO#/ \phantomsection\label{html:b161}
\hyperref[html:y161]{}
\item[links<sub>161</sub> =] BEI # term [links] \phantomsection\label{html:b300}
\hyperref[html:y300]{}
\item[quantifier<sub>300</sub> =] number /BOI#/<br /> | VEI # mex /VEhO#/ \phantomsection\label{html:b310}
\hyperref[html:y310]{}
\item[mex<sub>310</sub> =] mex-1 [operator mex-1] ...<br /> | FUhA # rp-expression \phantomsection\label{html:b311}
\hyperref[html:y311]{}
\item[mex-1<sub>311</sub> =] mex-2 [BIhE # operator mex-1] \phantomsection\label{html:b312}
\hyperref[html:y312]{}
\item[mex-2<sub>312</sub> =] operand<br /> | [PEhO #] operator mex-2 ... /KUhE#/ \phantomsection\label{html:b330}
\hyperref[html:y330]{}
\item[rp-expression<sub>330</sub> =] rp-operand rp-operand operator \phantomsection\label{html:b332}
\hyperref[html:y332]{}
\item[rp-operand<sub>332</sub> =] operand<br /> | rp-expression \phantomsection\label{html:b370}
\hyperref[html:y370]{}
\item[operator<sub>370</sub> =] operator-1 [joik-jek operator-1 | joik [stag] KE # operator /KEhE#/] ... \phantomsection\label{html:b371}
\hyperref[html:y371]{}
\item[operator-1<sub>371</sub> =] operator-2<br /> | guhek operator-1 gik operator-2<br /> | operator-2 (jek | joik) [stag] BO # operator-1 \phantomsection\label{html:b372}
\hyperref[html:y372]{}
\item[operator-2<sub>372</sub> =] mex-operator<br /> | KE # operator /KEhE#/ \phantomsection\label{html:b374}
\hyperref[html:y374]{}
\item[mex-operator<sub>374</sub> =] SE # mex-operator<br /> | NAhE # mex-operator<br /> | MAhO # mex /TEhU#/<br /> | NAhU # selbri /TEhU#/<br /> | VUhU # \phantomsection\label{html:b381}
\hyperref[html:y381]{}
\item[operand<sub>381</sub> =] operand-1 [(ek | joik) [stag] KE # operand /KEhE#/] \phantomsection\label{html:b382}
\hyperref[html:y382]{}
\item[operand-1<sub>382</sub> =] operand-2 [joik-ek operand-2] ... \phantomsection\label{html:b383}
\hyperref[html:y383]{}
\item[operand-2<sub>383</sub> =] operand-3 [(ek | joik) [stag] BO # operand-2] \phantomsection\label{html:b385}
\hyperref[html:y385]{}
\item[operand-3<sub>385</sub> =] quantifier<br /> | lerfu-string /BOI#/<br /> | NIhE # selbri /TEhU#/<br /> | MOhE # sumti /TEhU#/<br /> | JOhI # mex-2 ... /TEhU#/<br /> | gek operand gik operand-3<br /> | (LAhE # | NAhE BO #) operand /LUhU#/ \phantomsection\label{html:b812}
\hyperref[html:y812]{}
\item[number<sub>812</sub> =] PA [PA | lerfu-word] ... \phantomsection\label{html:b817}
\hyperref[html:y817]{}
\item[lerfu-string<sub>817</sub> =] lerfu-word [PA | lerfu-word] ... \phantomsection\label{html:b987}
\hyperref[html:y987]{}
\item[lerfu-word<sub>987</sub> =] BY<br /> | any-word BU<br /> | LAU lerfu-word<br /> | TEI lerfu-string FOI \phantomsection\label{html:b802}
\hyperref[html:y802]{}
\item[ek<sub>802</sub> =] [NA] [SE] A [NAI] \phantomsection\label{html:b818}
\hyperref[html:y818]{}
\item[gihek<sub>818</sub> =] [NA] [SE] GIhA [NAI] \phantomsection\label{html:b805}
\hyperref[html:y805]{}
\item[jek<sub>805</sub> =] [NA] [SE] JA [NAI] \phantomsection\label{html:b806}
\hyperref[html:y806]{}
\item[joik<sub>806</sub> =] [SE] JOI [NAI]<br /> | interval<br /> | GAhO interval GAhO \phantomsection\label{html:b932}
\hyperref[html:y932]{}
\item[interval<sub>932</sub> =] [SE] BIhI [NAI] \phantomsection\label{html:b421}
\hyperref[html:y421]{}
\item[joik-ek<sub>421</sub> =] joik #<br /> | ek # \phantomsection\label{html:b422}
\hyperref[html:y422]{}
\item[joik-jek<sub>422</sub> =] joik #<br /> | jek # \phantomsection\label{html:b807}
\hyperref[html:y807]{}
\item[gek<sub>807</sub> =] [SE] GA [NAI] #<br /> | joik GI #<br /> | stag gik \phantomsection\label{html:b808}
\hyperref[html:y808]{}
\item[guhek<sub>808</sub> =] [SE] GUhA [NAI] # \phantomsection\label{html:b816}
\hyperref[html:y816]{}
\item[gik<sub>816</sub> =] GI [NAI] # \phantomsection\label{html:b491}
\hyperref[html:y491]{}
\item[tag<sub>491</sub> =] tense-modal [joik-jek tense-modal] ... \phantomsection\label{html:b971}
\hyperref[html:y971]{}
\item[stag<sub>971</sub> =] simple-tense-modal [(jek | joik) simple-tense-modal] ... \phantomsection\label{html:b815}
\hyperref[html:y815]{}
\item[tense-modal<sub>815</sub> =] simple-tense-modal #<br /> | FIhO # selbri /FEhU#/ \phantomsection\label{html:b972}
\hyperref[html:y972]{}
\item[simple-tense-modal<sub>972</sub> =] [NAhE] [SE] BAI [NAI] [KI]<br /> | [NAhE] (time [space]<br /> | space [time]) \&CAhA [KI]<br /> | KI<br /> | CUhE \phantomsection\label{html:b1030}
\hyperref[html:y1030]{}
\item[time<sub>1030</sub> =] ZI \&time-offset ... \&ZEhA [PU [NAI]] \&      interval-property ... \phantomsection\label{html:b1033}
\hyperref[html:y1033]{}
\item[time-offset<sub>1033</sub> =] PU [NAI] [ZI] \phantomsection\label{html:b1040}
\hyperref[html:y1040]{}
\item[space<sub>1040</sub> =] VA \&space-offset ... \&space-interval \&      (MOhI space-offset) \phantomsection\label{html:b1045}
\hyperref[html:y1045]{}
\item[space-offset<sub>1045</sub> =] FAhA [NAI] [VA] \phantomsection\label{html:b1046}
\hyperref[html:y1046]{}
\item[space-interval<sub>1046</sub> =] ((VEhA \&VIhA) [FAhA [NAI]]) \&space-int-props \phantomsection\label{html:b1049}
\hyperref[html:y1049]{}
\item[space-int-props<sub>1049</sub> =] (FEhE interval-property) ... \phantomsection\label{html:b1051}
\hyperref[html:y1051]{}
\item[interval-property<sub>1051</sub> =] number ROI [NAI]<br /> | TAhE [NAI]<br /> | ZAhO [NAI] \phantomsection\label{html:b32}
\hyperref[html:y32]{}
\item[free<sub>32</sub> =] SEI # [terms [CU #]] selbri /SEhU/<br /> | SOI # sumti [sumti] /SEhU/<br /> | vocative [relative-clauses] selbri [relative-clauses] /DOhU/<br /> | vocative [relative-clauses] CMENE ... # [relative-clauses] /DOhU/<br /> | vocative [sumti] /DOhU/<br /> | (number | lerfu-string) MAI<br /> | TO text /TOI/<br /> | XI # (number | lerfu-string) /BOI/<br /> | XI # VEI # mex /VEhO/ \phantomsection\label{html:b415}
\hyperref[html:y415]{}
\item[vocative<sub>415</sub> =] (COI [NAI]) ... \&DOI \phantomsection\label{html:b411}
\hyperref[html:y411]{}
\item[indicators<sub>411</sub> =] [FUhE] indicator ... \phantomsection\label{html:b413}
\hyperref[html:y413]{}
\item[indicator<sub>413</sub> =] (UI | CAI) [NAI]<br /> | Y<br /> | DAhO<br /> | FUhO
\end{description}

The following rules are non-formal:
\begin{description}
      <dd>\phantomsection\label{html:b1100}
\hyperref[html:y1100]{}</dd>
\item[word<sub>1100</sub> =] [BAhE] any-word [indicators]
\item[any-word =] \q{any single word (no compound cmavo)}
\item[anything =] \q{any text at all, whether Lojban or not} \phantomsection\label{html:b1101}
\hyperref[html:y1101]{}\item[null<sub>1101</sub> =] any-word SI<br /> | utterance SA<br /> | text SU
\end{description}

FAhO is a universal terminator and signals the end of parsable
    input.



\sect{EBNF Cross-Reference}
\begin{description}
\item[A] \hyperref[html:b802]{ek}<sub>802</sub>
\item[BAI] \hyperref[html:b972]{simple-tense-modal}<sub>972</sub>
\item[BAhE] \hyperref[html:b1100]{word}<sub>1100</sub>
\item[BE] \hyperref[html:b160]{linkargs}<sub>160</sub>
\item[BEI] \hyperref[html:b161]{links}<sub>161</sub>
\item[BEhO] \hyperref[html:b160]{linkargs}<sub>160</sub>
\item[BIhE] \hyperref[html:b311]{mex-1}<sub>311</sub>
\item[BIhI] \hyperref[html:b932]{interval}<sub>932</sub>
\item[BO] \hyperref[html:b52]{bridi-tail-2}<sub>52</sub>, \hyperref[html:b383]{operand-2}<sub>383</sub>, \hyperref[html:b385]{operand-3}<sub>385</sub>, \hyperref[html:b371]{operator-1}<sub>371</sub>, \hyperref[html:b135]{selbri-5}<sub>135</sub>, \hyperref[html:b136]{selbri-6}<sub>136</sub>, \hyperref[html:b13]{statement-2}<sub>13</sub>, \hyperref[html:b93]{sumti-3}<sub>93</sub>, \hyperref[html:b97]{sumti-6}<sub>97</sub>, \hyperref[html:b2]{text-1}<sub>2</sub>
\item[BOI] \hyperref[html:b32]{free}<sub>32</sub>, \hyperref[html:b385]{operand-3}<sub>385</sub>, \hyperref[html:b300]{quantifier}<sub>300</sub>, \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[BRIVLA] \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[BU] \hyperref[html:b987]{lerfu-word}<sub>987</sub>
\item[BY] \hyperref[html:b987]{lerfu-word}<sub>987</sub>
\item[CAI] \hyperref[html:b413]{indicator}<sub>413</sub>
\item[CAhA] \hyperref[html:b972]{simple-tense-modal}<sub>972</sub>
\item[CEI] \hyperref[html:b150]{tanru-unit}<sub>150</sub>
\item[CEhE] \hyperref[html:b82]{terms-2}<sub>82</sub>
\item[CMENE] \hyperref[html:b32]{free}<sub>32</sub>, \hyperref[html:b97]{sumti-6}<sub>97</sub>, \hyperref[html:b0]{text}<sub>0</sub>
\item[CO] \hyperref[html:b132]{selbri-2}<sub>132</sub>
\item[COI] \hyperref[html:b415]{vocative}<sub>415</sub>
\item[CU] \hyperref[html:b32]{free}<sub>32</sub>, \hyperref[html:b40]{sentence}<sub>40</sub>
\item[CUhE] \hyperref[html:b972]{simple-tense-modal}<sub>972</sub>
\item[DAhO] \hyperref[html:b413]{indicator}<sub>413</sub>
\item[DOI] \hyperref[html:b415]{vocative}<sub>415</sub>
\item[DOhU] \hyperref[html:b32]{free}<sub>32</sub>
\item[FA] \hyperref[html:b83]{term}<sub>83</sub>
\item[FAhA] \hyperref[html:b1046]{space-interval}<sub>1046</sub>, \hyperref[html:b1045]{space-offset}<sub>1045</sub>
\item[FEhE] \hyperref[html:b1049]{space-int-props}<sub>1049</sub>
\item[FEhU] \hyperref[html:b815]{tense-modal}<sub>815</sub>
\item[FIhO] \hyperref[html:b815]{tense-modal}<sub>815</sub>
\item[FOI] \hyperref[html:b987]{lerfu-word}<sub>987</sub>
\item[FUhA] \hyperref[html:b310]{mex}<sub>310</sub>
\item[FUhE] \hyperref[html:b411]{indicators}<sub>411</sub>
\item[FUhO] \hyperref[html:b413]{indicator}<sub>413</sub>
\item[GA] \hyperref[html:b807]{gek}<sub>807</sub>
\item[GAhO] \hyperref[html:b806]{joik}<sub>806</sub>
\item[GEhU] \hyperref[html:b122]{relative-clause}<sub>122</sub>
\item[GI] \hyperref[html:b807]{gek}<sub>807</sub>, \hyperref[html:b816]{gik}<sub>816</sub>
\item[GIhA] \hyperref[html:b818]{gihek}<sub>818</sub>
\item[GOI] \hyperref[html:b122]{relative-clause}<sub>122</sub>
\item[GOhA] \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[GUhA] \hyperref[html:b808]{guhek}<sub>808</sub>
\item[I] \hyperref[html:b10]{paragraph}<sub>10</sub>, \hyperref[html:b12]{statement-1}<sub>12</sub>, \hyperref[html:b13]{statement-2}<sub>13</sub>, \hyperref[html:b2]{text-1}<sub>2</sub>
\item[JA] \hyperref[html:b805]{jek}<sub>805</sub>
\item[JAI] \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[JOI] \hyperref[html:b806]{joik}<sub>806</sub>
\item[JOhI] \hyperref[html:b385]{operand-3}<sub>385</sub>
\item[KE] \hyperref[html:b50]{bridi-tail}<sub>50</sub>, \hyperref[html:b54]{gek-sentence}<sub>54</sub>, \hyperref[html:b381]{operand}<sub>381</sub>, \hyperref[html:b372]{operator-2}<sub>372</sub>, \hyperref[html:b370]{operator}<sub>370</sub>, \hyperref[html:b134]{selbri-4}<sub>134</sub>, \hyperref[html:b91]{sumti-1}<sub>91</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[KEI] \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[KEhE] \hyperref[html:b50]{bridi-tail}<sub>50</sub>, \hyperref[html:b54]{gek-sentence}<sub>54</sub>, \hyperref[html:b381]{operand}<sub>381</sub>, \hyperref[html:b372]{operator-2}<sub>372</sub>, \hyperref[html:b370]{operator}<sub>370</sub>, \hyperref[html:b134]{selbri-4}<sub>134</sub>, \hyperref[html:b91]{sumti-1}<sub>91</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[KI] \hyperref[html:b972]{simple-tense-modal}<sub>972</sub>
\item[KOhA] \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[KU] \hyperref[html:b95]{sumti-5}<sub>95</sub>, \hyperref[html:b97]{sumti-6}<sub>97</sub>, \hyperref[html:b83]{term}<sub>83</sub>
\item[KUhE] \hyperref[html:b312]{mex-2}<sub>312</sub>
\item[KUhO] \hyperref[html:b122]{relative-clause}<sub>122</sub>
\item[LA] \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[LAU] \hyperref[html:b987]{lerfu-word}<sub>987</sub>
\item[LAhE] \hyperref[html:b385]{operand-3}<sub>385</sub>, \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[LE] \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[LEhU] \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[LI] \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[LIhU] \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[LOhO] \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[LOhU] \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[LU] \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[LUhU] \hyperref[html:b385]{operand-3}<sub>385</sub>, \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[MAI] \hyperref[html:b32]{free}<sub>32</sub>
\item[MAhO] \hyperref[html:b374]{mex-operator}<sub>374</sub>
\item[ME] \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[MEhU] \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[MOI] \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[MOhE] \hyperref[html:b385]{operand-3}<sub>385</sub>
\item[MOhI] \hyperref[html:b1040]{space}<sub>1040</sub>
\item[NA] \hyperref[html:b802]{ek}<sub>802</sub>, \hyperref[html:b20]{fragment}<sub>20</sub>, \hyperref[html:b54]{gek-sentence}<sub>54</sub>, \hyperref[html:b818]{gihek}<sub>818</sub>, \hyperref[html:b805]{jek}<sub>805</sub>, \hyperref[html:b131]{selbri-1}<sub>131</sub>, \hyperref[html:b83]{term}<sub>83</sub>
\item[NAI] \hyperref[html:b802]{ek}<sub>802</sub>, \hyperref[html:b807]{gek}<sub>807</sub>, \hyperref[html:b818]{gihek}<sub>818</sub>, \hyperref[html:b816]{gik}<sub>816</sub>, \hyperref[html:b808]{guhek}<sub>808</sub>, \hyperref[html:b413]{indicator}<sub>413</sub>, \hyperref[html:b1051]{interval-property}<sub>1051</sub>, \hyperref[html:b932]{interval}<sub>932</sub>, \hyperref[html:b805]{jek}<sub>805</sub>, \hyperref[html:b806]{joik}<sub>806</sub>, \hyperref[html:b972]{simple-tense-modal}<sub>972</sub>, \hyperref[html:b1046]{space-interval}<sub>1046</sub>, \hyperref[html:b1045]{space-offset}<sub>1045</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>, \hyperref[html:b0]{text}<sub>0</sub>, \hyperref[html:b1033]{time-offset}<sub>1033</sub>, \hyperref[html:b1030]{time}<sub>1030</sub>, \hyperref[html:b415]{vocative}<sub>415</sub>
\item[NAhE] \hyperref[html:b374]{mex-operator}<sub>374</sub>, \hyperref[html:b385]{operand-3}<sub>385</sub>, \hyperref[html:b136]{selbri-6}<sub>136</sub>, \hyperref[html:b972]{simple-tense-modal}<sub>972</sub>, \hyperref[html:b97]{sumti-6}<sub>97</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[NAhU] \hyperref[html:b374]{mex-operator}<sub>374</sub>
\item[NIhE] \hyperref[html:b385]{operand-3}<sub>385</sub>
\item[NIhO] \hyperref[html:b4]{paragraphs}<sub>4</sub>, \hyperref[html:b2]{text-1}<sub>2</sub>
\item[NOI] \hyperref[html:b122]{relative-clause}<sub>122</sub>
\item[NU] \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[NUhA] \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[NUhI] \hyperref[html:b85]{termset}<sub>85</sub>
\item[NUhU] \hyperref[html:b85]{termset}<sub>85</sub>
\item[PA] \hyperref[html:b817]{lerfu-string}<sub>817</sub>, \hyperref[html:b812]{number}<sub>812</sub>
\item[PEhE] \hyperref[html:b81]{terms-1}<sub>81</sub>
\item[PEhO] \hyperref[html:b312]{mex-2}<sub>312</sub>
\item[PU] \hyperref[html:b1033]{time-offset}<sub>1033</sub>, \hyperref[html:b1030]{time}<sub>1030</sub>
\item[RAhO] \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[ROI] \hyperref[html:b1051]{interval-property}<sub>1051</sub>
\item[SA] \hyperref[html:b1101]{null}<sub>1101</sub>
\item[SE] \hyperref[html:b802]{ek}<sub>802</sub>, \hyperref[html:b807]{gek}<sub>807</sub>, \hyperref[html:b818]{gihek}<sub>818</sub>, \hyperref[html:b808]{guhek}<sub>808</sub>, \hyperref[html:b932]{interval}<sub>932</sub>, \hyperref[html:b805]{jek}<sub>805</sub>, \hyperref[html:b806]{joik}<sub>806</sub>, \hyperref[html:b374]{mex-operator}<sub>374</sub>, \hyperref[html:b972]{simple-tense-modal}<sub>972</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[SEI] \hyperref[html:b32]{free}<sub>32</sub>
\item[SEhU] \hyperref[html:b32]{free}<sub>32</sub>
\item[SI] \hyperref[html:b1101]{null}<sub>1101</sub>
\item[SOI] \hyperref[html:b32]{free}<sub>32</sub>
\item[SU] \hyperref[html:b1101]{null}<sub>1101</sub>
\item[TAhE] \hyperref[html:b1051]{interval-property}<sub>1051</sub>
\item[TEI] \hyperref[html:b987]{lerfu-word}<sub>987</sub>
\item[TEhU] \hyperref[html:b374]{mex-operator}<sub>374</sub>, \hyperref[html:b385]{operand-3}<sub>385</sub>
\item[TO] \hyperref[html:b32]{free}<sub>32</sub>
\item[TOI] \hyperref[html:b32]{free}<sub>32</sub>
\item[TUhE] \hyperref[html:b14]{statement-3}<sub>14</sub>
\item[TUhU] \hyperref[html:b14]{statement-3}<sub>14</sub>
\item[UI] \hyperref[html:b413]{indicator}<sub>413</sub>
\item[VA] \hyperref[html:b1045]{space-offset}<sub>1045</sub>, \hyperref[html:b1040]{space}<sub>1040</sub>
\item[VAU] \hyperref[html:b20]{fragment}<sub>20</sub>, \hyperref[html:b71]{tail-terms}<sub>71</sub>
\item[VEI] \hyperref[html:b32]{free}<sub>32</sub>, \hyperref[html:b300]{quantifier}<sub>300</sub>
\item[VEhA] \hyperref[html:b1046]{space-interval}<sub>1046</sub>
\item[VEhO] \hyperref[html:b32]{free}<sub>32</sub>, \hyperref[html:b300]{quantifier}<sub>300</sub>
\item[VIhA] \hyperref[html:b1046]{space-interval}<sub>1046</sub>
\item[VUhO] \hyperref[html:b90]{sumti}<sub>90</sub>
\item[VUhU] \hyperref[html:b374]{mex-operator}<sub>374</sub>
\item[XI] \hyperref[html:b32]{free}<sub>32</sub>
\item[Y] \hyperref[html:b413]{indicator}<sub>413</sub>
\item[ZAhO] \hyperref[html:b1051]{interval-property}<sub>1051</sub>
\item[ZEI] \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[ZEhA] \hyperref[html:b1030]{time}<sub>1030</sub>
\item[ZI] \hyperref[html:b1033]{time-offset}<sub>1033</sub>, \hyperref[html:b1030]{time}<sub>1030</sub>
\item[ZIhE] \hyperref[html:b121]{relative-clauses}<sub>121</sub>
\item[ZO] \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[ZOI] \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[ZOhU] \hyperref[html:b30]{prenex}<sub>30</sub>
\item[any-word] \hyperref[html:b987]{lerfu-word}<sub>987</sub>, \hyperref[html:b1101]{null}<sub>1101</sub>, \hyperref[html:b97]{sumti-6}<sub>97</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>, \hyperref[html:b1100]{word}<sub>1100</sub>
\item[anything] \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[bridi-tail] \hyperref[html:b50]{bridi-tail}<sub>50</sub>, \hyperref[html:b40]{sentence}<sub>40</sub>
\item[bridi-tail-1] \hyperref[html:b50]{bridi-tail}<sub>50</sub>
\item[bridi-tail-2] \hyperref[html:b51]{bridi-tail-1}<sub>51</sub>, \hyperref[html:b52]{bridi-tail-2}<sub>52</sub>
\item[bridi-tail-3] \hyperref[html:b52]{bridi-tail-2}<sub>52</sub>
\item[ek] \hyperref[html:b20]{fragment}<sub>20</sub>, \hyperref[html:b421]{joik-ek}<sub>421</sub>, \hyperref[html:b383]{operand-2}<sub>383</sub>, \hyperref[html:b381]{operand}<sub>381</sub>, \hyperref[html:b91]{sumti-1}<sub>91</sub>, \hyperref[html:b93]{sumti-3}<sub>93</sub>
\item[fragment] \hyperref[html:b10]{paragraph}<sub>10</sub>
\item[free] \hyperref[html:b0]{text}<sub>0</sub>
\item[gek] \hyperref[html:b54]{gek-sentence}<sub>54</sub>, \hyperref[html:b385]{operand-3}<sub>385</sub>, \hyperref[html:b94]{sumti-4}<sub>94</sub>, \hyperref[html:b85]{termset}<sub>85</sub>
\item[gek-sentence] \hyperref[html:b53]{bridi-tail-3}<sub>53</sub>, \hyperref[html:b54]{gek-sentence}<sub>54</sub>
\item[gihek] \hyperref[html:b51]{bridi-tail-1}<sub>51</sub>, \hyperref[html:b52]{bridi-tail-2}<sub>52</sub>, \hyperref[html:b50]{bridi-tail}<sub>50</sub>, \hyperref[html:b20]{fragment}<sub>20</sub>
\item[gik] \hyperref[html:b54]{gek-sentence}<sub>54</sub>, \hyperref[html:b807]{gek}<sub>807</sub>, \hyperref[html:b385]{operand-3}<sub>385</sub>, \hyperref[html:b371]{operator-1}<sub>371</sub>, \hyperref[html:b136]{selbri-6}<sub>136</sub>, \hyperref[html:b94]{sumti-4}<sub>94</sub>, \hyperref[html:b85]{termset}<sub>85</sub>
\item[guhek] \hyperref[html:b371]{operator-1}<sub>371</sub>, \hyperref[html:b136]{selbri-6}<sub>136</sub>
\item[indicator] \hyperref[html:b411]{indicators}<sub>411</sub>
\item[indicators] \hyperref[html:b0]{text}<sub>0</sub>, \hyperref[html:b1100]{word}<sub>1100</sub>
\item[interval] \hyperref[html:b806]{joik}<sub>806</sub>
\item[interval-property] \hyperref[html:b1049]{space-int-props}<sub>1049</sub>, \hyperref[html:b1030]{time}<sub>1030</sub>
\item[jek] \hyperref[html:b422]{joik-jek}<sub>422</sub>, \hyperref[html:b371]{operator-1}<sub>371</sub>, \hyperref[html:b135]{selbri-5}<sub>135</sub>, \hyperref[html:b971]{stag}<sub>971</sub>, \hyperref[html:b13]{statement-2}<sub>13</sub>, \hyperref[html:b2]{text-1}<sub>2</sub>
\item[joik] \hyperref[html:b807]{gek}<sub>807</sub>, \hyperref[html:b421]{joik-ek}<sub>421</sub>, \hyperref[html:b422]{joik-jek}<sub>422</sub>, \hyperref[html:b383]{operand-2}<sub>383</sub>, \hyperref[html:b381]{operand}<sub>381</sub>, \hyperref[html:b371]{operator-1}<sub>371</sub>, \hyperref[html:b370]{operator}<sub>370</sub>, \hyperref[html:b134]{selbri-4}<sub>134</sub>, \hyperref[html:b135]{selbri-5}<sub>135</sub>, \hyperref[html:b971]{stag}<sub>971</sub>, \hyperref[html:b13]{statement-2}<sub>13</sub>, \hyperref[html:b91]{sumti-1}<sub>91</sub>, \hyperref[html:b93]{sumti-3}<sub>93</sub>, \hyperref[html:b2]{text-1}<sub>2</sub>
\item[joik-ek] \hyperref[html:b382]{operand-1}<sub>382</sub>, \hyperref[html:b92]{sumti-2}<sub>92</sub>
\item[joik-jek] \hyperref[html:b370]{operator}<sub>370</sub>, \hyperref[html:b134]{selbri-4}<sub>134</sub>, \hyperref[html:b12]{statement-1}<sub>12</sub>, \hyperref[html:b491]{tag}<sub>491</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>, \hyperref[html:b81]{terms-1}<sub>81</sub>, \hyperref[html:b0]{text}<sub>0</sub>
\item[lerfu-string] \hyperref[html:b32]{free}<sub>32</sub>, \hyperref[html:b987]{lerfu-word}<sub>987</sub>, \hyperref[html:b385]{operand-3}<sub>385</sub>, \hyperref[html:b97]{sumti-6}<sub>97</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[lerfu-word] \hyperref[html:b817]{lerfu-string}<sub>817</sub>, \hyperref[html:b987]{lerfu-word}<sub>987</sub>, \hyperref[html:b812]{number}<sub>812</sub>
\item[linkargs] \hyperref[html:b20]{fragment}<sub>20</sub>, \hyperref[html:b151]{tanru-unit-1}<sub>151</sub>
\item[links] \hyperref[html:b20]{fragment}<sub>20</sub>, \hyperref[html:b160]{linkargs}<sub>160</sub>, \hyperref[html:b161]{links}<sub>161</sub>
\item[mex] \hyperref[html:b32]{free}<sub>32</sub>, \hyperref[html:b374]{mex-operator}<sub>374</sub>, \hyperref[html:b300]{quantifier}<sub>300</sub>, \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[mex-1] \hyperref[html:b311]{mex-1}<sub>311</sub>, \hyperref[html:b310]{mex}<sub>310</sub>
\item[mex-2] \hyperref[html:b311]{mex-1}<sub>311</sub>, \hyperref[html:b312]{mex-2}<sub>312</sub>, \hyperref[html:b385]{operand-3}<sub>385</sub>
\item[mex-operator] \hyperref[html:b374]{mex-operator}<sub>374</sub>, \hyperref[html:b372]{operator-2}<sub>372</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[number] \hyperref[html:b32]{free}<sub>32</sub>, \hyperref[html:b1051]{interval-property}<sub>1051</sub>, \hyperref[html:b300]{quantifier}<sub>300</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[operand] \hyperref[html:b312]{mex-2}<sub>312</sub>, \hyperref[html:b385]{operand-3}<sub>385</sub>, \hyperref[html:b381]{operand}<sub>381</sub>, \hyperref[html:b332]{rp-operand}<sub>332</sub>
\item[operand-1] \hyperref[html:b381]{operand}<sub>381</sub>
\item[operand-2] \hyperref[html:b382]{operand-1}<sub>382</sub>, \hyperref[html:b383]{operand-2}<sub>383</sub>
\item[operand-3] \hyperref[html:b383]{operand-2}<sub>383</sub>, \hyperref[html:b385]{operand-3}<sub>385</sub>
\item[operator] \hyperref[html:b311]{mex-1}<sub>311</sub>, \hyperref[html:b312]{mex-2}<sub>312</sub>, \hyperref[html:b310]{mex}<sub>310</sub>, \hyperref[html:b372]{operator-2}<sub>372</sub>, \hyperref[html:b370]{operator}<sub>370</sub>, \hyperref[html:b330]{rp-expression}<sub>330</sub>
\item[operator-1] \hyperref[html:b371]{operator-1}<sub>371</sub>, \hyperref[html:b370]{operator}<sub>370</sub>
\item[operator-2] \hyperref[html:b371]{operator-1}<sub>371</sub>
\item[paragraph] \hyperref[html:b4]{paragraphs}<sub>4</sub>
\item[paragraphs] \hyperref[html:b4]{paragraphs}<sub>4</sub>, \hyperref[html:b2]{text-1}<sub>2</sub>
\item[prenex] \hyperref[html:b20]{fragment}<sub>20</sub>, \hyperref[html:b11]{statement}<sub>11</sub>, \hyperref[html:b41]{subsentence}<sub>41</sub>
\item[quantifier] \hyperref[html:b20]{fragment}<sub>20</sub>, \hyperref[html:b385]{operand-3}<sub>385</sub>, \hyperref[html:b95]{sumti-5}<sub>95</sub>, \hyperref[html:b112]{sumti-tail-1}<sub>112</sub>
\item[relative-clause] \hyperref[html:b121]{relative-clauses}<sub>121</sub>
\item[relative-clauses] \hyperref[html:b20]{fragment}<sub>20</sub>, \hyperref[html:b32]{free}<sub>32</sub>, \hyperref[html:b95]{sumti-5}<sub>95</sub>, \hyperref[html:b97]{sumti-6}<sub>97</sub>, \hyperref[html:b112]{sumti-tail-1}<sub>112</sub>, \hyperref[html:b111]{sumti-tail}<sub>111</sub>, \hyperref[html:b90]{sumti}<sub>90</sub>
\item[rp-expression] \hyperref[html:b310]{mex}<sub>310</sub>, \hyperref[html:b332]{rp-operand}<sub>332</sub>
\item[rp-operand] \hyperref[html:b330]{rp-expression}<sub>330</sub>
\item[selbri] \hyperref[html:b53]{bridi-tail-3}<sub>53</sub>, \hyperref[html:b32]{free}<sub>32</sub>, \hyperref[html:b374]{mex-operator}<sub>374</sub>, \hyperref[html:b385]{operand-3}<sub>385</sub>, \hyperref[html:b131]{selbri-1}<sub>131</sub>, \hyperref[html:b136]{selbri-6}<sub>136</sub>, \hyperref[html:b95]{sumti-5}<sub>95</sub>, \hyperref[html:b112]{sumti-tail-1}<sub>112</sub>, \hyperref[html:b815]{tense-modal}<sub>815</sub>
\item[selbri-1] \hyperref[html:b130]{selbri}<sub>130</sub>
\item[selbri-2] \hyperref[html:b131]{selbri-1}<sub>131</sub>, \hyperref[html:b132]{selbri-2}<sub>132</sub>
\item[selbri-3] \hyperref[html:b132]{selbri-2}<sub>132</sub>, \hyperref[html:b134]{selbri-4}<sub>134</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[selbri-4] \hyperref[html:b133]{selbri-3}<sub>133</sub>
\item[selbri-5] \hyperref[html:b134]{selbri-4}<sub>134</sub>, \hyperref[html:b135]{selbri-5}<sub>135</sub>
\item[selbri-6] \hyperref[html:b135]{selbri-5}<sub>135</sub>, \hyperref[html:b136]{selbri-6}<sub>136</sub>
\item[sentence] \hyperref[html:b14]{statement-3}<sub>14</sub>, \hyperref[html:b41]{subsentence}<sub>41</sub>
\item[simple-tense-modal] \hyperref[html:b971]{stag}<sub>971</sub>, \hyperref[html:b815]{tense-modal}<sub>815</sub>
\item[space] \hyperref[html:b972]{simple-tense-modal}<sub>972</sub>
\item[space-int-props] \hyperref[html:b1046]{space-interval}<sub>1046</sub>
\item[space-interval] \hyperref[html:b1040]{space}<sub>1040</sub>
\item[space-offset] \hyperref[html:b1040]{space}<sub>1040</sub>
\item[stag] \hyperref[html:b52]{bridi-tail-2}<sub>52</sub>, \hyperref[html:b50]{bridi-tail}<sub>50</sub>, \hyperref[html:b807]{gek}<sub>807</sub>, \hyperref[html:b383]{operand-2}<sub>383</sub>, \hyperref[html:b381]{operand}<sub>381</sub>, \hyperref[html:b371]{operator-1}<sub>371</sub>, \hyperref[html:b370]{operator}<sub>370</sub>, \hyperref[html:b134]{selbri-4}<sub>134</sub>, \hyperref[html:b135]{selbri-5}<sub>135</sub>, \hyperref[html:b13]{statement-2}<sub>13</sub>, \hyperref[html:b91]{sumti-1}<sub>91</sub>, \hyperref[html:b93]{sumti-3}<sub>93</sub>, \hyperref[html:b2]{text-1}<sub>2</sub>
\item[statement] \hyperref[html:b10]{paragraph}<sub>10</sub>, \hyperref[html:b11]{statement}<sub>11</sub>
\item[statement-1] \hyperref[html:b11]{statement}<sub>11</sub>
\item[statement-2] \hyperref[html:b12]{statement-1}<sub>12</sub>, \hyperref[html:b13]{statement-2}<sub>13</sub>
\item[statement-3] \hyperref[html:b13]{statement-2}<sub>13</sub>
\item[subsentence] \hyperref[html:b54]{gek-sentence}<sub>54</sub>, \hyperref[html:b122]{relative-clause}<sub>122</sub>, \hyperref[html:b41]{subsentence}<sub>41</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[sumti] \hyperref[html:b32]{free}<sub>32</sub>, \hyperref[html:b385]{operand-3}<sub>385</sub>, \hyperref[html:b91]{sumti-1}<sub>91</sub>, \hyperref[html:b94]{sumti-4}<sub>94</sub>, \hyperref[html:b97]{sumti-6}<sub>97</sub>, \hyperref[html:b112]{sumti-tail-1}<sub>112</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>, \hyperref[html:b83]{term}<sub>83</sub>
\item[sumti-1] \hyperref[html:b90]{sumti}<sub>90</sub>
\item[sumti-2] \hyperref[html:b91]{sumti-1}<sub>91</sub>
\item[sumti-3] \hyperref[html:b92]{sumti-2}<sub>92</sub>, \hyperref[html:b93]{sumti-3}<sub>93</sub>
\item[sumti-4] \hyperref[html:b93]{sumti-3}<sub>93</sub>, \hyperref[html:b94]{sumti-4}<sub>94</sub>
\item[sumti-5] \hyperref[html:b94]{sumti-4}<sub>94</sub>
\item[sumti-6] \hyperref[html:b95]{sumti-5}<sub>95</sub>, \hyperref[html:b111]{sumti-tail}<sub>111</sub>
\item[sumti-tail] \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[sumti-tail-1] \hyperref[html:b111]{sumti-tail}<sub>111</sub>
\item[tag] \hyperref[html:b54]{gek-sentence}<sub>54</sub>, \hyperref[html:b130]{selbri}<sub>130</sub>, \hyperref[html:b14]{statement-3}<sub>14</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>, \hyperref[html:b83]{term}<sub>83</sub>
\item[tail-terms] \hyperref[html:b51]{bridi-tail-1}<sub>51</sub>, \hyperref[html:b52]{bridi-tail-2}<sub>52</sub>, \hyperref[html:b53]{bridi-tail-3}<sub>53</sub>, \hyperref[html:b50]{bridi-tail}<sub>50</sub>, \hyperref[html:b54]{gek-sentence}<sub>54</sub>
\item[tanru-unit] \hyperref[html:b136]{selbri-6}<sub>136</sub>
\item[tanru-unit-1] \hyperref[html:b150]{tanru-unit}<sub>150</sub>
\item[tanru-unit-2] \hyperref[html:b151]{tanru-unit-1}<sub>151</sub>, \hyperref[html:b152]{tanru-unit-2}<sub>152</sub>
\item[tense-modal] \hyperref[html:b491]{tag}<sub>491</sub>
\item[term] \hyperref[html:b160]{linkargs}<sub>160</sub>, \hyperref[html:b161]{links}<sub>161</sub>, \hyperref[html:b122]{relative-clause}<sub>122</sub>, \hyperref[html:b82]{terms-2}<sub>82</sub>
\item[terms] \hyperref[html:b20]{fragment}<sub>20</sub>, \hyperref[html:b32]{free}<sub>32</sub>, \hyperref[html:b30]{prenex}<sub>30</sub>, \hyperref[html:b40]{sentence}<sub>40</sub>, \hyperref[html:b71]{tail-terms}<sub>71</sub>, \hyperref[html:b85]{termset}<sub>85</sub>
\item[terms-1] \hyperref[html:b80]{terms}<sub>80</sub>
\item[terms-2] \hyperref[html:b81]{terms-1}<sub>81</sub>
\item[termset] \hyperref[html:b83]{term}<sub>83</sub>
\item[text] \hyperref[html:b32]{free}<sub>32</sub>, \hyperref[html:b1101]{null}<sub>1101</sub>, \hyperref[html:b97]{sumti-6}<sub>97</sub>
\item[text-1] \hyperref[html:b14]{statement-3}<sub>14</sub>, \hyperref[html:b0]{text}<sub>0</sub>
\item[time] \hyperref[html:b972]{simple-tense-modal}<sub>972</sub>
\item[time-offset] \hyperref[html:b1030]{time}<sub>1030</sub>
\item[utterance] \hyperref[html:b1101]{null}<sub>1101</sub>
\item[vocative] \hyperref[html:b32]{free}<sub>32</sub>
\end{description}
